---
title: "03 вЂ” Construct structure & measurement invariance"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

# The question

What latent structure organizes the nine core tasks вЂ” and can we trust that
structure across **age** and **site**? Picking a factor model from the data is
deceptively hard here because three things are entangled:

1. **The structure changes developmentally** (factors differentiate with age).
2. **Sites may not share the structure** (different languages, populations,
   administration), and BogotГЎ's partial battery leaves some task pairs with no
   joint coverage.
3. **Single indicators** вЂ” most LEVANTE constructs are one task, so a task's
   *reliable specific* variance is confounded with *error*; we cannot even ask
   "is ToM separable?" without separating the two.

The discipline that organizes this: **measurement before structure.** "3 factors
vs 2," "factors differentiate with age," "BogotГЎ differs" can all be
*measurement non-invariance masquerading as structure*. So we (a) disattenuate
to handle the single-indicator problem, (b) select a structure, then (c) test
whether its **measurement model is invariant** across site and age before
trusting any structural claim. (Notebook 04 complements this with the
exploratory ESEM and the smooth age-differentiation curves; here we are
confirmatory.)

*(The previous version of this notebook вЂ” construct-level latent growth curves вЂ”
is preserved at `old/03_sem_growth_lgc.qmd`.)*

```{r setup}
source(here::here("common.R"))
library(lavaan)
library(tidySEM)
sc <- read_rds(here("data/scores_all_sites.rds"))
sites4 <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota",
            "pilot_uniandes_co_rural","pilot_western_ca_main")
core <- c("matrix","mrot","math","hf","mg","sds","trog","vocab","tom")
fitm <- function(f, m) as.numeric(fitMeasures(f, m))
heywood <- function(f){ pe <- parameterEstimates(f)
  any(pe$est[pe$op=="~~" & pe$lhs==pe$rhs] < -1e-4) ||
    any(abs(subset(standardizedSolution(f), op=="=~")$est.std) > 1.001) }
```

```{r data}
fo <- sc |> filter(dataset %in% sites4, task_id %in% core_task_ids, !is.na(score)) |>
  group_by(user_id, short) |> arrange(age) |> slice(1) |> ungroup()        # first obs per childГ—task
rel <- fo |> group_by(short) |>
  summarise(rel = var(score)/(var(score) + mean(score_se^2, na.rm = TRUE)), .groups = "drop") |>
  filter(short %in% core) |> deframe()
age1 <- fo |> group_by(user_id) |> arrange(age) |> slice(1) |> ungroup() |> select(user_id, age)
xw <- fo |> filter(short %in% core) |> group_by(short) |> mutate(z = as.numeric(scale(score))) |> ungroup() |>
  select(user_id, dataset, short, z) |> distinct(user_id, short, .keep_all = TRUE) |>
  pivot_wider(names_from = short, values_from = z) |>
  left_join(age1, by = "user_id") |> mutate(age_c = age - mean(age, na.rm = TRUE))
```

Pooled cross-sectional sample (first observation per child, four data-rich
sites): **N = `r nrow(xw)`**. Per-task marginal reliabilities (from the EAP
SEs) range `r round(min(rel),2)`вЂ“`r round(max(rel),2)`.

# 1. The single-indicator problem: disattenuated task correlations

To ask whether any task carries reliable variance *distinct* from the others,
we first strip measurement error. Each task becomes a **single-indicator latent
with its error variance fixed** from its reliability
(`eta_t =~ 1*t; t ~~ (1в€’ПЃ_t)В·t`, scores standardized), and we read off the
error-corrected correlations. We use Germany (the complete-battery site) to
avoid coverage gaps.

```{r disattenuated}
de <- xw |> filter(dataset == "pilot_mpieva_de_main")
silv <- paste(c(sprintf("eta_%s =~ 1*%s", core, core),
                sprintf("%s ~~ %.4f*%s", core, pmax(1 - rel[core], .02), core)), collapse = "\n")
f_cor <- cfa(silv, de, missing = "fiml")
cm <- lavInspect(f_cor, "cor.lv"); dimnames(cm) <- list(core, core)

cor_heatmap <- function(cm, title){
  m <- cm; diag(m) <- NA
  df <- as.data.frame(m) |> rownames_to_column("t1") |> pivot_longer(-t1, names_to = "t2", values_to = "r")
  mid <- mean(range(df$r, na.rm = TRUE))
  df |> mutate(t1 = factor(t1, levels = rev(core)), t2 = factor(t2, levels = core), dark = r > mid) |>
    ggplot(aes(t2, t1, fill = r)) + geom_tile(color = "white") +
    geom_text(aes(label = ifelse(is.na(r), "", sprintf("%.2f", r)), color = dark), size = 3, show.legend = FALSE) +
    scale_fill_viridis_c(na.value = "grey92", name = "r") +
    scale_color_manual(values = c(`TRUE` = "grey10", `FALSE` = "white"), na.value = "grey40") +
    coord_fixed() + labs(x = NULL, y = NULL, title = title) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
}
cor_heatmap(cm, "Disattenuated (error-corrected) task correlations вЂ” Germany")
```

Two things this settles immediately:

- **ToM is *not* separable.** Error-corrected, it correlates в‰€ 0.90 with TROG
  and в‰€ 0.86 with vocab вЂ” its reliable variance is essentially the
  verbal/language cluster, leaving little room for an independent "social
  cognition" construct at the task level.
- **The most distinct task is Hearts & Flowers** (lowest correlations, в‰€ 0.3вЂ“0.6)
  вЂ” the inhibition/EF task carries the most reliable variance not shared with the
  rest. Even so, every off-diagonal is positive and large: a strong positive
  manifold.

# 2. Which structure? (confirmatory comparison)

We compare a single general factor (**g**), the emergent **2-factor**
(fluid / verbal), the theory **3-factor** (reasoning / EF / language), and a
**bifactor** (g + orthogonal domain specifics) вЂ” standard CFA, pooled
(FIML), age partialled. Because a strong positive manifold makes fit indices
keep nudging toward more (near-collinear, often improper) factors, we read the
table by **propriety first**, then parsimony (BIC), then fit.

```{r structures}
mods <- list(
  `g`        = "g =~ matrix+mrot+math+hf+mg+sds+trog+vocab+tom\n g ~ age_c",
  `2-factor` = "fluid =~ matrix+mrot+math+hf+mg+sds\n verbal =~ trog+vocab+tom\n fluid ~ age_c; verbal ~ age_c",
  `3-factor` = "reasoning =~ matrix+mrot+math\n ef =~ hf+mg+sds\n language =~ trog+vocab+tom\n reasoning~age_c; ef~age_c; language~age_c",
  `bifactor` = "g =~ matrix+mrot+math+hf+mg+sds+trog+vocab+tom\n reasoning =~ matrix+mrot+math\n ef =~ hf+mg+sds\n language =~ trog+vocab+tom\n g ~ age_c")
fits <- imap(mods, ~cfa(.x, xw, std.lv = TRUE, missing = "fiml", orthogonal = (.y == "bifactor")))
map_dfr(fits, ~tibble(CFI = fitm(.x,"cfi"), RMSEA = fitm(.x,"rmsea"), BIC = fitm(.x,"bic"),
                      proper = !heywood(.x)), .id = "model") |>
  mutate(across(c(CFI,RMSEA), \(x) round(x,3)), BIC = round(BIC)) |>
  knitr::kable(caption = "Structure comparison (pooled, age-partialled). proper = no Heywood case.")
```

```{r f3-cor}
lavInspect(fits[["3-factor"]], "cor.lv")[c("reasoning","ef","language"), c("reasoning","ef","language")] |>
  round(2) |> knitr::kable(caption = "3-factor inter-factor correlations (pooled)")
```

The **bifactor is improper** (a Heywood case вЂ” its specific factors
collapse/explode), itself a finding: *beyond g, domain-specific variance is
negligible.* Among the **proper** models the **2-factor (fluid/verbal) is
BIC-preferred** (BIC 25638 vs 3-factor 25657 vs g 25708) and ties the 3-factor on
fit (CFI в‰€ 0.95) вЂ” the third factor does not earn its parsimony cost. And the
3-factor's factors are near-collinear (reasoningвЂ“EF в‰€ 0.96). So the honest read
is **a dominant general factor with ~two weakly-separable domains** ("g plus
modest domains"), consistent with 04's ESEM. We carry the proper candidates
(g / 2-factor / 3-factor) into the invariance tests.

The three proper candidates as path diagrams (standardized loadings, pooled
cross-sectional). The curved arrows between latent factors show how strongly the
domains correlate вЂ” visibly near-collinear in the 2- and 3-factor models.

```{r sem-diagrams, fig.width=10, fig.height=3.6}
diag_mods <- list(
  "1-factor (g)" = "g =~ matrix+mrot+math+hf+mg+sds+trog+vocab+tom",
  "2-factor (fluid / verbal)" = "fluid =~ matrix+mrot+math+hf+mg+sds\n verbal =~ trog+vocab+tom",
  "3-factor (reasoning / EF / language)" = "reasoning =~ matrix+mrot+math\n ef =~ hf+mg+sds\n language =~ trog+vocab+tom")
for (nm in names(diag_mods)) {
  f <- cfa(diag_mods[[nm]], xw, std.lv = TRUE, missing = "fiml")
  print(graph_sem(f, variance_diameter = .15, text_size = 2.6, fix_coord = TRUE) +
        labs(title = nm) +
        theme(panel.background = element_rect(fill = "white", color = NA),
              plot.title = element_text(size = 11)))
}
```

# 3. Does the structure replicate across sites?

A full all-site multigroup is blocked by BogotГЎ's partial battery (zero joint
coverage of some task pairs в†’ FIML breaks). So we (a) fit each model **separately
per site** (descriptive replication, incl. Canada), and (b) run a **formal
2-group multigroup** for the two sites with full mutual coverage (Germany +
Canada).

```{r per-site}
meas <- list(g = "g =~ matrix+mrot+math+hf+mg+sds+trog+vocab+tom",
             `2-factor` = "fluid =~ matrix+mrot+math+hf+mg+sds\n verbal =~ trog+vocab+tom",
             `3-factor` = "reasoning =~ matrix+mrot+math\n ef =~ hf+mg+sds\n language =~ trog+vocab+tom")
grp <- xw |> mutate(site = recode(dataset, pilot_mpieva_de_main="Germany",
  pilot_uniandes_co_bogota="BogotГЎ-urban", pilot_uniandes_co_rural="BogotГЎ-rural", pilot_western_ca_main="Canada"))
per_site <- map_dfr(unique(grp$site), function(s){
  d <- filter(grp, site == s)
  map_dfr(meas, function(m){ f <- tryCatch(cfa(m, d, std.lv=TRUE, missing="fiml"), error=function(e) NULL)
    if(is.null(f)||!lavInspect(f,"converged")) return(tibble(CFI=NA, BIC=NA, max_factor_r=NA))
    cr <- tryCatch(lavInspect(f,"cor.lv"), error=function(e) NULL)
    tibble(CFI=fitm(f,"cfi"), BIC=fitm(f,"bic"),
           max_factor_r = if(!is.null(cr) && is.matrix(cr) && ncol(cr)>1) max(cr[upper.tri(cr)]) else NA_real_) }, .id="model") |>
    mutate(site = s, n = nrow(d)) })
per_site |> mutate(model = factor(model, names(meas))) |> arrange(site, model) |>
  select(site, n, model, CFI, BIC, max_factor_r) |>
  mutate(across(c(CFI,max_factor_r), \(x) round(x,3)), BIC = round(BIC)) |>
  knitr::kable(caption = "Per-site CFA (max_factor_r = strongest inter-factor correlation): both 2- and 3-factor beat g; factors near-collinear")
```

```{r mg-de-ca}
dc <- filter(grp, site %in% c("Germany","Canada"))
mg <- imap_dfr(meas[c("2-factor","3-factor")], function(m, modnm){
  map_dfr(list(configural=NULL, metric="loadings", scalar=c("loadings","intercepts")), function(eq){
    f <- tryCatch(cfa(m, dc, group="site", group.equal=eq, std.lv=TRUE, missing="fiml"), error=function(e) NULL)
    if(is.null(f)||!lavInspect(f,"converged")) return(tibble(CFI=NA, RMSEA=NA))
    tibble(CFI=fitm(f,"cfi"), RMSEA=fitm(f,"rmsea")) }, .id="invariance") |>
    mutate(model = modnm, dCFI = c(NA, round(diff(CFI),3))) })
mg |> mutate(across(c(CFI,RMSEA), \(x) round(x,3))) |> select(model, invariance, CFI, RMSEA, dCFI) |>
  knitr::kable(caption = "Site invariance ladders (Germany + Canada), 2- and 3-factor")
```

A multi-factor structure **replicates** in every site вЂ” both the 2- and 3-factor
beat g on BIC (CFI в‰€ 0.97вЂ“0.99). Notably the **2-factor stays proper everywhere**
(strongest factor r в‰€ 0.83вЂ“0.85), whereas the **3-factor goes improper in
BogotГЎ-rural** (reasoningвЂ“EF = 1.10 > 1 вЂ” the two fluid factors are empirically
indistinguishable there): another mark in the 2-factor's favour. The
Germany + Canada multigroup supports **configural invariance** for both
(CFI в‰€ 0.99), only **approximate metric** invariance (О”CFI в‰€ в€’0.029 for the
2-factor, в€’0.017 for the 3-factor вЂ” loadings roughly, not strictly, shared, and a
touch less so for the broader 2-factor), and **scalar fails decisively**
(О”CFI в‰€ в€’0.13). So across two very different sites the *form* is shared and
loadings approximately so, but **absolute score levels are not comparable** (and
we don't interpret them).

# 4. Does the structure hold across age?

This is the crucial test for the 04 differentiation finding. If factor
correlations change with age (differentiation) but **loadings stay invariant**,
the differentiation is *structural*. If loadings themselves drift with age, the
"differentiation" could be a measurement artifact. We test metric/scalar
invariance across age bins (pooled).

```{r age-invariance}
xa <- xw |> mutate(agebin = cut(age, c(0,8,10,20), labels = c("<8","8вЂ“10","10+")))
age_ladder <- function(m, lab) map_dfr(list(configural=NULL, metric="loadings", scalar=c("loadings","intercepts")), function(eq){
  f <- tryCatch(cfa(m, xa, group="agebin", group.equal=eq, std.lv=TRUE, missing="fiml"), error=function(e) NULL)
  if(is.null(f)||!lavInspect(f,"converged")) return(tibble(CFI=NA))
  tibble(CFI=fitm(f,"cfi")) }, .id="invariance") |> mutate(model=lab, dCFI=c(NA, round(diff(CFI),3)))
bind_rows(age_ladder(meas[["3-factor"]], "3-factor"),
          age_ladder("fluid =~ matrix+mrot+math+hf+mg+sds\n verbal =~ trog+vocab+tom", "2-factor")) |>
  mutate(CFI = round(CFI,3)) |> select(model, invariance, CFI, dCFI) |>
  knitr::kable(caption = "Age-bin invariance (<8 / 8вЂ“10 / 10+): loadings ~invariant, intercepts shift")
```

**Metric (loading) invariance approximately holds across age** (О”CFI в‰€ в€’0.01
configuralв†’metric) вЂ” the loadings barely move. **Scalar fails** (О”CFI в‰€ в€’0.03):
intercepts rise with age, exactly as expected when children improve. The
takeaway: the age-differentiation in 04 (factor *correlations* falling with age)
is **structural change, not loading drift** вЂ” the measurement model is stable
enough across age for that interpretation to stand. (See 04 for the smooth
local-SEM correlation curves under 2/3/4-factor structures.)

# Synthesis

```{r save}
write_rds(list(disattenuated = cm, structure_fits = map(fits, ~fitm(.x, c("cfi","rmsea","bic"))),
               per_site = per_site, site_invariance = mg),
          here::here("data/structure_invariance_03.rds"))
```

- **Single indicators:** disattenuating (fixed-error latents) shows the core
  tasks form a strong positive manifold; **ToM is not separable** from the
  verbal cluster (error-corrected r в‰€ 0.9), while **H&F** carries the most
  distinct reliable variance.
- **Structure:** a **dominant general factor with ~two weakly-separable
  domains.** Among proper models the **2-factor (fluid/verbal) is BIC-preferred**,
  tying the 3-factor on fit and staying **proper in every site** (the 3-factor
  goes improper in BogotГЎ-rural, reasoningвЂ“EF > 1); the bifactor's specifics
  degenerate. Consistent with 04's ESEM.
- **Site:** the multi-factor structure **replicates** across Germany, BogotГЎ, and
  Canada; the Germany + Canada multigroup supports configural and **approximate**
  metric invariance (О”CFI в‰€ в€’0.02 to в€’0.03 вЂ” form and roughly loadings shared) вЂ”
  but **not scalar** (levels not comparable).
- **Age:** loadings are **~invariant across age** (metric О”CFI в‰€ в€’0.01) while
  intercepts shift вЂ” so 04's age-differentiation is a genuine **structural**
  change, not a measurement artifact.

**Net:** the appropriate model is *g-with-~two-modest-domains*, invariant in form
and approximately in loadings across site and age, with comparability limited at
the level of absolute means (scalar invariance fails for both site and age).
Single-task constructs like ToM cannot be separated at the task level вЂ” testing
their specific variance would need the item-level upgrade flagged in the plan.
