---
title: "04 — Construct structure: g-loading, differentiation, and differential validity"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

# The question

03 found the core tasks resolve into a **strong general factor** with only
weakly-separable reasoning / EF / language domains (inter-factor r ≈ 0.85–0.96).
That's good for a single ability score but raises a worry: **little differential
validity** — the domains barely behave as distinct constructs. This notebook
interrogates that from four angles:

1. **Developmental differentiation** — do the factors separate with age (local
   SEM)?
2. **Empirical structure (ESEM)** — what factor solution do the data actually
   prefer, with cross-loadings allowed?
3. **Predictive differential validity** — does a domain factor predict a
   matching external criterion (reading) over and above the others?
4. **A shared-speed (method) artifact?** — does a general response-speed factor
   account for the cross-task covariation?

```{r setup}
source(here::here("common.R"))
library(lavaan); library(psych)
sc <- read_rds(here("data/scores_all_sites.rds"))
sites4 <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota",
            "pilot_uniandes_co_rural","pilot_western_ca_main")

fo <- sc |> filter(dataset %in% sites4, task_id %in% c(core_task_ids,"swr","sre","pa"),
                   !is.na(score)) |>
  group_by(user_id, short) |> arrange(age) |> slice(1) |> ungroup()
age1 <- fo |> group_by(user_id) |> arrange(age) |> slice(1) |> ungroup() |> select(user_id, age, dataset)
xw <- fo |> select(user_id, short, score) |> distinct(user_id, short, .keep_all = TRUE) |>
  pivot_wider(names_from = short, values_from = score) |>
  left_join(age1, by = "user_id")

meas <- "reasoning =~ matrix + mrot + math
         ef =~ hf + mg + sds
         language =~ trog + vocab + tom"
fcor <- function(fit){ c <- lavInspect(fit,"cor.lv")
  c(`R–EF`=c["reasoning","ef"], `R–Lang`=c["reasoning","language"], `EF–Lang`=c["ef","language"]) }
```

Pooled cross-sectional sample (first observation per child, four data-rich
sites): **N = `r nrow(xw)`**.

# 1. Developmental differentiation (local SEM)

We fit the 3-factor model at a sliding series of focal ages, weighting children
by a Gaussian kernel on age (bandwidth 1.5 yr), and read off the inter-factor
correlations as a smooth function of age. If abilities differentiate with
development, the correlations should fall with age.

```{r losem, cache=TRUE}
h <- 1.5
losem <- map_dfr(seq(6, 12, by = 0.5), function(a0) {
  w <- dnorm((xw$age - a0) / h)
  fit <- tryCatch(cfa(meas, cbind(xw, .w = w), std.lv = TRUE, missing = "fiml",
                      sampling.weights = ".w"), error = function(e) NULL)
  if (is.null(fit) || !lavInspect(fit, "converged")) return(NULL)
  bind_cols(focal_age = a0, eff_n = round(sum(w)), as_tibble_row(fcor(fit)))
})
losem |> pivot_longer(c(`R–EF`,`R–Lang`,`EF–Lang`), names_to="pair", values_to="r") |>
  ggplot(aes(focal_age, r, color = pair)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  scale_x_continuous(breaks = 6:12) + ylim(0.5, 1) +
  labs(x = "Focal age (years)", y = "Inter-factor correlation",
       title = "Abilities differentiate with age",
       subtitle = "Kernel-weighted (local) SEM, bandwidth 1.5 yr")
```

```{r losem-tab}
losem |> filter(focal_age %in% c(6,8,10,12)) |>
  mutate(across(c(`R–EF`,`R–Lang`,`EF–Lang`), \(x) round(x,2))) |>
  knitr::kable(caption = "Inter-factor correlations by focal age")
```

**Reasoning and EF stay fused (r ≈ 0.95→0.92) across the whole range, but
Language pulls away** — Reasoning–Language drops from ~0.95 to ~0.77 and
EF–Language from ~0.92 to ~0.66 between ages 6 and 12. This is the classic
*age-differentiation* pattern: the g-dominance is strongest in the youngest
children and partly dissolves as the verbal/language domain separates. So the
limited differential validity is, in part, **real developmental structure** —
not only a measurement artifact.

# 2. What structure do the data prefer? (ESEM)

Exploratory factor analysis on the FIML correlation matrix (handles the
partial-battery missingness), allowing items to cross-load.

```{r esem, cache=TRUE}
items9 <- xw |> select(matrix,mrot,math,hf,mg,sds,trog,vocab,tom)
R <- lavCor(items9, missing = "fiml"); N <- nrow(items9)
fit_tab <- map_dfr(1:4, function(nf){
  f <- suppressWarnings(fa(R, nfactors = nf, n.obs = N, fm = "ml", rotate = "oblimin"))
  tibble(factors = nf, RMSEA = f$RMSEA[1], TLI = f$TLI, BIC = f$BIC)
})
knitr::kable(fit_tab, digits = 3, caption = "EFA fit by number of factors (FIML correlation)")
```

```{r esem-loadings}
f3 <- suppressWarnings(fa(R, nfactors = 3, n.obs = N, fm = "ml", rotate = "oblimin"))
load3 <- unclass(f3$loadings); colnames(load3) <- paste0("F", 1:3)
as.data.frame(round(load3, 2)) |> rownames_to_column("task") |>
  knitr::kable(caption = "3-factor ESEM loadings (oblimin)")
round(f3$Phi, 2) |> knitr::kable(caption = "ESEM factor correlations")
```

The exploratory solution **does not recover reasoning vs. EF as separate
factors**. Instead:

- **F1 — a broad fluid/quantitative factor** loading matrix, mrot, math, mg,
  sds (≈ 0.66–0.83): reasoning *and* EF *and* math collapse together.
- **F2 — a language/verbal factor**: vocab (0.92), tom, trog.
- **F3** is weak and anchored by Hearts & Flowers (low-communality, the
  warm-up-heavy task).

F1 and F2 correlate ≈ 0.69. So empirically the battery is closer to **two
correlated factors — fluid-cognitive and verbal — than to the intended three**.
(BIC keeps improving past 3 factors, but the extra factors are small and
task-specific rather than substantive.) This matches §1: language is the one
domain that genuinely separates.

# 3. Predictive differential validity (reading criterion)

If the domains have differential validity, the **language** factor should
predict an external **reading** measure (ROAR word/sentence/phoneme) over and
above reasoning and EF. (MEFS, the EF criterion, is not in this data release, so
we can only test the reading side.)

```{r predval, cache=TRUE}
mod <- paste(meas,
  "reading =~ swr + sre + pa",
  "reading ~ reasoning + ef + language", sep = "\n")
fit_pv <- sem(mod, xw, std.lv = TRUE, missing = "fiml")
lavInspect(fit_pv, "cor.lv")["reading", c("reasoning","ef","language")] |> round(3) |>
  enframe("factor","cor_with_reading") |>
  knitr::kable(caption = "Latent correlation of each cognitive factor with the reading factor")
standardizedSolution(fit_pv) |> filter(lhs == "reading", op == "~") |>
  select(predictor = rhs, est.std, se, pvalue) |>
  knitr::kable(digits = 3, caption = "reading ~ reasoning + ef + language (note collinearity)")
```

**Reading correlates about equally with all three cognitive factors**
(reasoning 0.87, EF 0.81, language 0.85) — it is **not** preferentially related
to language. Reading is predicted by the *general* factor, not by a specific
language domain. (The regression coefficients are unstable from collinearity —
reasoning takes a large but high-SE coefficient — so the latent correlations are
the trustworthy read.) So even against an external criterion, **differential
validity is weak**: the cognitive factors relate to reading through g.

# 4. Is the g-loading a shared-speed (method) artifact?

All tasks share a format (4-AFC, tablet, one session). A general
response-speed/decision factor could inflate cross-task ability covariation. We
build a general-speed factor from per-person median (log) reaction times across
the 8 timed core tasks (Memory has no RT), standardized within task, and ask
whether it explains the cognitive g.

```{r rt, cache=TRUE}
rttasks <- setdiff(core_task_ids, "memory-game")
fr_runs <- sc |> filter(dataset %in% sites4, task_id %in% rttasks) |>
  group_by(user_id, task_id) |> arrange(age) |> slice(1) |> ungroup() |> pull(run_id)
trrt <- load_levante_trials(task_ids = rttasks) |>
  filter(run_id %in% fr_runs, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000) |>
  group_by(run_id, user_id, short) |>
  summarise(lrt = median(log(rt_numeric)), .groups = "drop")
rtwide <- trrt |> group_by(short) |> mutate(z = as.numeric(scale(lrt))) |> ungroup() |>
  select(user_id, short, z) |> distinct(user_id, short, .keep_all = TRUE) |>
  pivot_wider(names_from = short, values_from = z)
rtm <- rtwide |> select(-user_id) |> as.matrix()
cm <- cor(rtm, use = "pairwise")
mean_rt_r <- mean(cm[upper.tri(cm)], na.rm = TRUE)

gfit <- cfa("g =~ matrix+mrot+math+hf+mg+sds+trog+vocab+tom", xw, std.lv=TRUE, missing="fiml")
gsc <- tibble(user_id = xw$user_id, g = as.numeric(lavPredict(gfit)[, "g"]))
spd <- rtwide |> mutate(speed = rowMeans(across(-user_id), na.rm = TRUE)) |> select(user_id, speed)
sg <- gsc |> inner_join(spd, by = "user_id")
tibble(metric = c("mean cross-task RT correlation (general speed exists?)",
                  "cor(cognitive g, general slowness)"),
       value = c(round(mean_rt_r, 3), round(cor(sg$g, sg$speed, use="complete.obs"), 3))) |>
  knitr::kable(caption = "Speed factor and its relation to cognitive g")
```

A general speed factor *does* exist — reaction times correlate ≈ 0.34 across
tasks. But it is **nearly orthogonal to cognitive ability** (r(g, slowness) ≈
0.15, and *positive* — if anything, slower, more deliberate children score
slightly higher, consistent with rushing→errors elsewhere). So the cross-task
covariation in *ability* is **not** a shared-speed/method artifact; partialling
speed out would not differentiate the ability factors. The g-loading is
substantive, not procedural.

# Synthesis

Four angles converge on a coherent picture of the g-loading:

| Hypothesis | Verdict |
|---|---|
| Between-site mean differences inflate it (03 probe) | **No** — it's within-site |
| Shared response-speed / method factor | **No** — speed ≈ orthogonal to g (r 0.15) |
| Purely a measurement defect | **Partly not** — it's developmentally real |
| Reasoning/EF/Language are separable | **Only Language** separates (ESEM: 2 factors) |
| Domains have differential (criterion) validity | **Weak** — reading relates to g, not specifically language |

So the battery, at these ages, measures a **strong general ability plus a
separable verbal/language dimension**, and that verbal dimension **differentiates
from the fluid core as children get older**. The limited differential validity
is therefore partly a true developmental fact (young children's abilities are
less differentiated), not just an artifact — though it does mean that, for now,
a single g-score plus a language score captures most of what the battery
measures.

**Implications / next steps:**

- Report and model **age-moderated structure** (the differentiation curve) as a
  substantive result rather than treating high factor correlations as a flaw.
- For differential-validity claims, focus on the **older** end of the range and
  on the **fluid-vs-verbal** split (the one the data support), not a three-way
  reasoning/EF/language split.
- Bring in **MEFS** (when available) and other external criteria to test the EF
  side of differential validity — the reading side already shows g, not domain,
  prediction.
- A longitudinal second-order model (does *g* grow; does the structure
  differentiate within child over time?) is the natural follow-up as waves
  accrue.

```{r save}
write_rds(list(losem = losem, esem = f3, fit_pv = fit_pv,
               rt = list(mean_r = mean_rt_r, g_speed = cor(sg$g, sg$speed, use="complete.obs"))),
          here::here("data/construct_structure_04.rds"))
```
