---
title: "07 — RT intra-individual variability: purification and cross-task structure"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

# The question

In `06` the most promising within-child variability index was **RT
intra-individual variability (RT-IIV)** — but the *naive* per-run SD of log-RT
showed almost no cross-task coherence (mean r ≈ 0.05) and only ~4% stable-person
variance. Two things could be hiding a real signal: (a) the raw SD is
**contaminated** by systematic structure (within-task time-course,
autocorrelation, administration differences), and (b) RT-IIV is a second-moment
estimate, so it is **noisy when trial counts are low**.

This notebook does the cleanup the feedback asked for: purify the RT signal, then
re-examine whether **cross-task RT-IIV cohere** — both across all tasks and
restricted to tasks/runs with enough trials.

```{r setup}
source(here::here("common.R"))
library(splines); library(psych)
sc <- read_rds(here("data/scores_all_sites.rds"))
sites4 <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota",
            "pilot_uniandes_co_rural","pilot_western_ca_main")
rt_tasks <- tibble(
  task_id = c("egma-math","matrix-reasoning","mental-rotation","trog","vocab",
              "same-different-selection","hearts-and-flowers","theory-of-mind","swr","sre","pa"),
  short   = c("math","matrix","mrot","trog","vocab","sds","hf","tom","swr","sre","pa"))
adap <- c("egma-math","matrix-reasoning","mental-rotation","same-different-selection","trog","vocab")

# ---- analysis parameter: which trials feed RT-IIV (carried throughout) ----
USE_TRIALS <- "all"      # "all" or "correct"

runs <- sc |> filter(dataset %in% sites4, task_id %in% rt_tasks$task_id, !is.na(score)) |>
  filter(!(task_id %in% adap) | adaptive) |>
  group_by(user_id, task_id) |> arrange(age) |> slice(1) |> ungroup() |>
  select(run_id, user_id, task_id, age)
```

The trial-set choice (**`r USE_TRIALS`** trials) is a parameter, applied
everywhere below; a correct-only sensitivity check appears in §1.4.

# 1. Purifying the RT signal

**What actually needs removing.** IIV here is the SD of log-RT *within a run*.
Because the SD is location-invariant, any nuisance that is **constant within a
run** — the child's mean speed, the run's mean, a **site/administration
intercept** — drops out automatically (and in log-space a multiplicative
site/device slowdown is just such an additive shift). What inflates the
within-run SD is structure that **varies across trials within a run**: the
**trial-order time-course** (warm-up, fatigue) and **autocorrelation**. So those
are what we detrend and filter. Site/age, which can shift the *level* of a
child's IIV (administration medium, development), are removed at the **IIV
level** instead (§3).

## 1.1 The within-task time-course (detrending)

Log-RT drifts systematically over the course of a task. We fit a task-level
model `log RT ~ ns(trial_number, 4) + site` and work with the residuals.

```{r extract, cache=TRUE}
purify_task <- function(tid, use_trials){
  rn <- runs |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rn$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000,
           !is.na(trial_number)) |>
    left_join(rn |> select(run_id, age), by = "run_id")
  if (use_trials == "correct") tr <- tr |> filter(correct == 1)
  tr <- tr |> mutate(lrt = log(rt_numeric))
  nsite <- n_distinct(tr$dataset)
  form <- if (nsite > 1) lrt ~ ns(trial_number, 4) + dataset else lrt ~ ns(trial_number, 4)
  tr$resid <- resid(lm(form, data = tr))
  # AR(1) innovations within run (pooled rho per task)
  tr <- tr |> group_by(run_id) |> arrange(trial_number, .by_group = TRUE) |>
    mutate(lag_resid = lag(resid)) |> ungroup()
  rho <- cor(tr$resid, tr$lag_resid, use = "complete.obs")
  tr <- tr |> mutate(innov = resid - rho * lag_resid)
  trend <- tr |> mutate(tn = pmin(trial_number, quantile(trial_number, .95))) |>
    group_by(short = rt_tasks$short[rt_tasks$task_id == tid], tn) |>
    summarise(m = mean(lrt - mean(lrt)), .groups = "drop")
  per_run <- tr |> group_by(run_id, user_id) |>
    summarise(ntrial = n(),
              iiv_raw   = sd(lrt),
              iiv_detr  = sd(resid),
              iiv_innov = sd(innov, na.rm = TRUE), .groups = "drop") |>
    filter(ntrial >= 8) |>
    mutate(short = rt_tasks$short[rt_tasks$task_id == tid])
  list(short = rt_tasks$short[rt_tasks$task_id == tid], rho = rho,
       per_run = per_run, trend = trend, med_n = median(per_run$ntrial))
}
ex <- map(rt_tasks$task_id, purify_task, use_trials = USE_TRIALS)
iiv <- map_dfr(ex, "per_run")
trend <- map_dfr(ex, "trend")
```

```{r trend-plot, fig.height=4.3}
trend |> ggplot(aes(tn, m, color = short)) +
  geom_line(linewidth = 0.8) + scale_color_viridis_d(name = NULL) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  labs(x = "Trial number", y = "Mean log-RT (centered)",
       title = "Within-task time-course of log-RT (the trend we detrend out)",
       subtitle = "Most tasks slow slightly then drift; speeded tasks (swr/hf) speed up with warm-up")
```

## 1.2 Autocorrelation

Consecutive RTs are positively autocorrelated, so the raw SD mixes
trial-to-trial noise with slow drift. We estimate a per-task AR(1) coefficient
on the detrended residuals (within run) and also form the **innovations**
(uncorrelated component) as a more aggressive purification.

## 1.3 How much each step changes the IIV

```{r purify-summary}
sb <- function(a, b){ r <- cor(a, b, use = "complete.obs"); 2*r/(1+r) }
rel_one <- function(tid, use_trials){
  rn <- runs |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rn$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number)) |>
    mutate(lrt = log(rt_numeric))
  if (use_trials == "correct") tr <- tr |> filter(correct == 1)
  nsite <- n_distinct(tr$dataset); form <- if (nsite>1) lrt ~ ns(trial_number,4)+dataset else lrt ~ ns(trial_number,4)
  tr$resid <- resid(lm(form, data = tr))
  sh <- tr |> group_by(run_id) |> filter(n() >= 12) |> mutate(h = row_number() %% 2) |>
    group_by(run_id, h) |> summarise(v = sd(resid), .groups = "drop") |>
    pivot_wider(names_from = h, values_from = v, names_prefix = "h")
  sb(sh$h0, sh$h1)
}
summ <- map_dfr(seq_along(ex), function(i){
  e <- ex[[i]]
  tibble(short = e$short, median_trials = e$med_n, n_runs = nrow(e$per_run), ar1_rho = e$rho,
         `cor(raw,detrended)` = cor(e$per_run$iiv_raw, e$per_run$iiv_detr),
         `cor(detrended,AR-innov)` = cor(e$per_run$iiv_detr, e$per_run$iiv_innov),
         iiv_reliability = rel_one(rt_tasks$task_id[i], USE_TRIALS))
})
summ |> arrange(desc(iiv_reliability)) |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Purification diagnostics per task (AR(1) ρ, step effects, split-half reliability)")
```

Autocorrelation is substantial everywhere (ρ ≈ 0.3–0.6). Detrending moves the
IIV more for some tasks than others (`cor(raw, detrended)` 0.84–0.99), and the
AR-innovation IIV stays highly correlated with the detrended one (≥ 0.84), so the
two purification choices broadly agree. We carry the **detrended** IIV forward as
primary and use the AR-innovation as a robustness variant.

## 1.4 Sensitivity: all vs correct trials, and an age-varying trend

```{r sens-trials}
ex_corr <- map(rt_tasks$task_id, purify_task, use_trials = "correct")
iiv_corr <- map_dfr(ex_corr, "per_run") |> select(run_id, short, iiv_detr_c = iiv_detr)
sens <- iiv |> select(run_id, short, iiv_detr) |> inner_join(iiv_corr, by = c("run_id","short")) |>
  group_by(short) |> summarise(`cor(all, correct-only)` = cor(iiv_detr, iiv_detr_c), .groups = "drop")
sens |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "All-trials vs correct-only detrended IIV agree closely per task")
```

```{r sens-age}
# does the within-task trend vary by age? compare detrended IIV with vs without ns(trial)*age
age_check <- map_dfr(rt_tasks$task_id, function(tid){
  rn <- runs |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rn$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number)) |>
    left_join(rn |> select(run_id, age), by = "run_id") |> mutate(lrt = log(rt_numeric), age_c = age - 9) |>
    filter(!is.na(age_c))                       # keep complete cases so both models use the same rows
  if (USE_TRIALS == "correct") tr <- tr |> filter(correct == 1)
  r0 <- resid(lm(lrt ~ ns(trial_number,4) + dataset, data = tr))
  r1 <- resid(lm(lrt ~ ns(trial_number,4) * age_c + dataset, data = tr))
  pr <- tibble(run_id = tr$run_id, r0, r1) |> group_by(run_id) |>
    summarise(s0 = sd(r0), s1 = sd(r1), .groups = "drop")
  tibble(short = rt_tasks$short[rt_tasks$task_id == tid],
         `cor(no-age, age×trial)` = cor(pr$s0, pr$s1, use = "complete.obs"))
})
age_check |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Letting the trial-order trend vary with age barely changes the IIV (r ≈ 1)")
```

Two checks: across tasks the **all-trials and correct-only** IIVs are
moderately-to-highly correlated (r ≈ 0.74–0.98 — most similar for the high-trial
speeded tasks, least for H&F and math), so including error trials shifts the
estimate somewhat but largely preserves the rank ordering; and letting the
trial-order trend **vary with age** leaves the IIV essentially unchanged (r ≈ 1).
We default to the simpler detrend (`ns(trial) + site`) on **all** trials, with
the trial set carried as a parameter (`USE_TRIALS`) for easy flipping.

**Is "all trials" the more valid choice?** Before locking in the default, we
compare both trial sets on **internal reliability** (split-half) and
**longitudinal stability** (wave-1 vs wave-2, Leipzig + Bogotá) for *every* task.
Error trials (impulsive errors, lapses) may carry attentional-variability signal —
especially in speeded tasks — but for an untimed task a wrong answer to a hard
item is not a lapse, so the better choice could differ by task.

```{r trialset-alltasks, cache=TRUE}
long_sites <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota","pilot_uniandes_co_rural")
ts_metrics <- function(tid, ut){
  wv <- sc |> filter(dataset %in% long_sites, task_id == tid, !is.na(score)) |>
    group_by(user_id) |> arrange(age) |> mutate(wave = row_number(), nwave = max(wave)) |>
    filter(nwave >= 2, wave <= 2) |> ungroup() |> select(run_id, user_id, dataset, wave)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% wv$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number)) |>
    mutate(lrt = log(rt_numeric))
  if (ut == "correct") tr <- tr |> filter(correct == 1)
  if (nrow(tr) < 100) return(NULL)
  nsite <- n_distinct(tr$dataset); form <- if (nsite>1) lrt ~ ns(trial_number,4)+dataset else lrt ~ ns(trial_number,4)
  tr$resid <- resid(lm(form, data = tr))
  pr <- tr |> group_by(run_id) |> filter(n() >= 8) |> summarise(iiv = sd(resid), .groups = "drop") |> left_join(wv, by = "run_id")
  sh <- tr |> group_by(run_id) |> filter(n() >= 12) |> mutate(h = row_number() %% 2) |>
    group_by(run_id, h) |> summarise(v = sd(resid), .groups = "drop") |> pivot_wider(names_from = h, values_from = v, names_prefix = "h")
  rr <- cor(sh$h0, sh$h1, use = "complete.obs")
  w <- pr |> select(user_id, wave, iiv) |> pivot_wider(names_from = wave, values_from = iiv, names_prefix = "w") |> filter(!is.na(w1), !is.na(w2))
  tibble(short = rt_tasks$short[rt_tasks$task_id == tid], trials = ut,
         reliability = 2*rr/(1+rr), test_retest = cor(w$w1, w$w2), n_rt = nrow(w))
}
ts <- map_dfr(rt_tasks$task_id, \(t) bind_rows(ts_metrics(t, "all"), ts_metrics(t, "correct"))) |> filter(n_rt >= 30)
```

```{r trialset-table}
ts |> select(short, trials, reliability, test_retest) |>
  pivot_wider(names_from = trials, values_from = c(reliability, test_retest)) |>
  transmute(task = short, `reliability: all` = reliability_all, `reliability: correct` = reliability_correct,
            `test–retest: all` = test_retest_all, `test–retest: correct` = test_retest_correct) |>
  arrange(desc(`reliability: all`)) |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "All trials vs correct-only RT-IIV, per task: split-half reliability and longitudinal test–retest")
```

```{r trialset-plot, fig.height=4.3}
ord <- ts |> filter(trials == "all") |> arrange(reliability) |> pull(short)
ts |> pivot_longer(c(reliability, test_retest), names_to = "metric", values_to = "value") |>
  mutate(short = factor(short, levels = ord),
         metric = recode(metric, reliability = "split-half reliability", test_retest = "test–retest stability")) |>
  ggplot(aes(value, short, color = trials)) +
  geom_line(aes(group = interaction(short, metric)), color = "grey75") +
  geom_point(size = 2.4) + facet_wrap(~metric, scales = "free_x") +
  scale_color_manual(values = c(all = "#268bd2", correct = "#cb4b16"), name = "trials") +
  labs(x = NULL, y = NULL, title = "All trials vs correct-only: reliability and longitudinal stability",
       subtitle = "All trials wins reliability on nearly every task and stability on most") +
  theme(legend.position = "bottom")
```

Across tasks, **all trials is the better default**:

- It gives **higher split-half reliability on 9 of 11 tasks** — including every
  signal-carrying task — because including error trials adds trials and steadies
  the SD estimate. (The two exceptions, Same-&-Different and ROAR-Phoneme, are
  the smallest/noisiest samples.)
- It gives **higher test–retest stability on the majority (7 of 11)**, with the
  largest gains exactly where RT-IIV matters most — Hearts & Flowers (0.47 vs
  0.28), matrix, ROAR-Word, vocab. Correct-only nudges ahead on four
  low-stability tasks (math, TROG, Same-&-Different, ROAR-Sentence), but by small
  margins and where stability is near the floor anyway.

For a speeded executive task the logic is clearest — the impulsive errors and
lapses we'd discard *are* the attentional signal — but the all-trials advantage
holds broadly. We therefore use **all trials** throughout.

# 2. Reliability depends on trial count

```{r rel-ntrials, fig.height=4.3}
summ |> ggplot(aes(median_trials, iiv_reliability)) +
  geom_smooth(method = "lm", se = FALSE, color = "grey70", linewidth = 0.7) +
  geom_point(aes(size = n_runs), alpha = 0.7, color = "#268bd2") +
  ggrepel::geom_text_repel(aes(label = short), size = 3.2) +
  scale_x_continuous(trans = "log10") +
  labs(x = "Median trials per run (log scale)", y = "Split-half reliability of RT-IIV",
       size = "n runs", title = "RT-IIV reliability rises with trial count",
       subtitle = "swr (120 trials) ≈ 0.94; tasks under ~25 trials fall below ~0.55")
```

Reliability climbs steeply with trial count: ROAR-Word (120 trials) reaches
≈ 0.94, while the shortest tasks (pa, mental rotation, ~ 19–26 trials) sit
around 0.5. This is the core practical lesson — **RT-IIV is only a dependable
per-child measure when the task delivers enough trials** — and motivates the
trial-count–filtered analysis in §3.

# 3. Do cross-task RT-IIV cohere?

We build the child × task matrix of (detrended) RT-IIV — keeping site and age
variance in (we examine those as effects of interest in §4, not removing them) —
and examine the correlation matrix and factor structure, first across **all
tasks**, then restricted to tasks that clear a reliability/trial-count bar.

```{r crosstask}
iiv_adj <- iiv |> group_by(short) |> mutate(iiv_z = as.numeric(scale(iiv_detr))) |> ungroup()
wide <- iiv_adj |> select(user_id, short, iiv_z) |> distinct(user_id, short, .keep_all = TRUE) |>
  pivot_wider(names_from = short, values_from = iiv_z)
cmat <- cor(wide |> select(-user_id), use = "pairwise")
cor_heatmap <- function(cm, title){
  m <- cm; diag(m) <- NA
  df <- as.data.frame(m) |> rownames_to_column("t1") |> pivot_longer(-t1, names_to = "t2", values_to = "r")
  mid <- mean(range(df$r, na.rm = TRUE))
  df |> mutate(t1 = factor(t1, levels = rev(rownames(cm))), t2 = factor(t2, levels = rownames(cm)), dark = r > mid) |>
    ggplot(aes(t2, t1, fill = r)) + geom_tile(color = "white") +
    geom_text(aes(label = ifelse(is.na(r), "", sprintf("%.2f", r)), color = dark), size = 3, show.legend = FALSE) +
    scale_fill_viridis_c(na.value = "grey92", name = "r") +
    scale_color_manual(values = c(`TRUE` = "grey10", `FALSE` = "white"), na.value = "grey40") +
    coord_fixed() + labs(x = NULL, y = NULL, title = title) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
}
cor_heatmap(cmat, "Cross-task correlations of purified RT-IIV (all tasks)")
```

```{r fa-all}
ord <- rt_tasks$short
fa_report <- function(cm, N, label){
  fits <- map_dfr(1:3, function(nf){ f <- suppressWarnings(fa(cm, nfactors = nf, n.obs = N, fm = "ml", rotate = "oblimin"))
    tibble(factors = nf, TLI = f$TLI, RMSEA = f$RMSEA[1]) })
  list(fits = fits, load = suppressWarnings(fa(cm, nfactors = 2, n.obs = N, fm = "ml", rotate = "oblimin")))
}
mean_off <- function(cm) mean(cm[upper.tri(cm)], na.rm = TRUE)
allfa <- fa_report(cmat, nrow(wide), "all")
allfa$fits |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = glue::glue("EFA fit, all tasks (mean off-diagonal r = {round(mean_off(cmat),3)})"))
L <- unclass(allfa$load$loadings); colnames(L) <- paste0("F", 1:ncol(L))
as.data.frame(round(L, 2)) |> rownames_to_column("task") |>
  knitr::kable(caption = "2-factor loadings, all tasks")
```

```{r fa-filtered}
keep <- summ |> filter(iiv_reliability >= 0.6, median_trials >= 25, n_runs >= 150) |> pull(short)
widef <- iiv_adj |> filter(short %in% keep) |> select(user_id, short, iiv_z) |>
  distinct(user_id, short, .keep_all = TRUE) |> pivot_wider(names_from = short, values_from = iiv_z)
cmatf <- cor(widef |> select(-user_id), use = "pairwise")
cor_heatmap(cmatf, glue::glue("Reliable subset: {paste(keep, collapse=', ')} (mean off-diagonal r = {round(mean_off(cmatf),3)})"))
fff <- fa_report(cmatf, nrow(widef), "filtered")
fff$fits |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "EFA fit, reliable subset")
Lf <- unclass(suppressWarnings(fa(cmatf, nfactors = 1, n.obs = nrow(widef), fm = "ml"))$loadings)
as.data.frame(round(Lf, 2)) |> rownames_to_column("task") |> arrange(desc(ML1)) |>
  knitr::kable(caption = "1-factor loadings, reliable subset")
```

The picture is **weak, two-stranded coherence — not one general trait**:

- Purification roughly doubles the cross-task correlation versus the naive
  per-run SD (notebook 06's ≈ 0.05 → ≈ 0.10). A single general RT-IIV factor still
  does **not** fit (1-factor TLI ≈ 0.38).
- The 2-factor solution shows a **clear speeded-task factor** (ROAR-Word .93,
  ROAR-Sentence .55, Hearts & Flowers .54 — where rapid responding makes RT-IIV a
  clean attention/lapse signal) and a **weaker untimed-cognitive strand** (matrix
  .40, mental rotation .38, TROG .33, math .28 — RT driven by problem
  difficulty/strategy). The second factor is partly inflated by the small-sample
  Same-&-Different task (n ≈ 78), which drops out under the reliability filter.
- Restricting to the **reliable, high-trial tasks** (matrix, vocab, H&F, ToM,
  ROAR-Word, ROAR-Sentence) raises the mean correlation to ≈ 0.17 and yields a
  passable single factor (1-factor TLI ≈ 0.84) **dominated by the speeded tasks**
  (ROAR-Word .99, H&F .51, ROAR-Sentence .41, ToM .38), with the cognitive tasks
  weak (matrix .18, vocab .10) — consistent with the naive null being partly
  attenuation from noisy, low-trial estimates, the durable signal living in the
  speeded tasks.

# 4. Site and age differences in RT-IIV

Rather than scrubbing site and age out of the IIV, we examine them directly: do
children respond more variably at some **sites** (administration medium, setting,
population) or at younger **ages**?

```{r site-age-prep}
library(lme4)
iiv_sa <- iiv |> left_join(runs |> select(run_id, age), by = "run_id") |>
  left_join(sc |> distinct(run_id, dataset), by = "run_id") |>
  mutate(site = recode(dataset,
           pilot_mpieva_de_main = "DE (Leipzig)", pilot_uniandes_co_bogota = "CO (Bogotá-urban)",
           pilot_uniandes_co_rural = "CO (Bogotá-rural)", pilot_western_ca_main = "CA (Western)"),
         iiv_z = ave(iiv_detr, short, FUN = \(x) as.numeric(scale(x))), age_c = age - 9)
```

## 4.1 Age

```{r age-plot, fig.height=4.3}
slope <- fixef(lmer(iiv_detr ~ age + (1|short), data = iiv_sa))["age"]
iiv_sa |> ggplot(aes(age, iiv_detr, color = short)) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.7) +
  scale_color_viridis_d(name = NULL) +
  labs(x = "Age (years)", y = "Purified RT-IIV (SD detrended log-RT)",
       title = glue::glue("Purified RT-IIV vs age (pooled slope {round(slope,3)} log-RT-SD/yr)"),
       subtitle = "Modest overall decline — concentrated in the speeded tasks")
```

```{r age-slopes}
iiv_sa |> group_by(short) |> summarise(age_slope = coef(lm(iiv_detr ~ age))[2], .groups = "drop") |>
  arrange(age_slope) |> mutate(age_slope = round(age_slope, 4)) |>
  knitr::kable(caption = "Per-task age slope of RT-IIV (negative = more consistent with age)")
```

RT-IIV declines with age overall, but the decline is **concentrated in the
speeded tasks** — ROAR-Word (−0.051/yr), ROAR-Sentence (−0.029), Hearts & Flowers
(−0.027) — exactly the tasks that carry the cross-task RT-IIV signal (§3), while
the untimed cognitive tasks are essentially flat. Developmental gains in
response consistency show up where RT-IIV is a clean measure.

## 4.2 Site

```{r site-means, fig.height=4.3}
iiv_sa |> group_by(short, site) |> summarise(m = mean(iiv_z), se = sd(iiv_z)/sqrt(n()), .groups = "drop") |>
  ggplot(aes(site, m, color = site)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
  geom_pointrange(aes(ymin = m - se, ymax = m + se), size = 0.3) +
  facet_wrap(~short, nrow = 2) + scale_color_brewer(palette = "Set1", guide = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  labs(x = NULL, y = "RT-IIV (within-task z)", title = "RT-IIV by site, per task",
       subtitle = "Bogotá (urban & rural) consistently above Leipzig / Western")
```

```{r site-model}
m_site <- lmer(iiv_z ~ dataset + age_c + (1|short), data = iiv_sa)
vc <- as.data.frame(VarCorr(lmer(iiv_z ~ (1|short) + (1|dataset), data = iiv_sa)))
co <- summary(m_site)$coefficients
tibble(term = c("CO Bogotá-urban vs DE","CO Bogotá-rural vs DE","CA Western vs DE","age (per year)"),
       estimate = co[c("datasetpilot_uniandes_co_bogota","datasetpilot_uniandes_co_rural",
                       "datasetpilot_western_ca_main","age_c"), "Estimate"],
       t = co[c("datasetpilot_uniandes_co_bogota","datasetpilot_uniandes_co_rural",
                "datasetpilot_western_ca_main","age_c"), "t value"]) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = glue::glue(
    "Site & age effects on RT-IIV (z), age-adjusted; site explains {round(vc$vcov[vc$grp=='dataset']/sum(vc$vcov)*100,1)}% of total IIV variance"))
```

There are clear **group-level site differences**: children in Bogotá — both
urban (+0.28 SD) and rural (+0.33 SD) — are substantially **more RT-variable**
than Leipzig even after age adjustment, while Western ≈ Leipzig. These are large,
consistent mean shifts, though site accounts for only ~4% of *total* RT-IIV
variance (individual differences dominate). The cause is not identifiable here —
it could be administration medium (devices, supervision, setting), population, or
language — but it is a clear caution against treating RT-IIV as directly
comparable across sites without accounting for it, and a flag for the
measurement-invariance of variability indices.

# 5. Longitudinal stability

Split-half reliability (§2) is a *within-session* property — it says nothing
about whether a child's RT-IIV is stable over time. The real trait test is
**test–retest stability** across waves. We use the longitudinal samples
(**Leipzig** + **Bogotá** urban & rural), take each child's first two
administrations per task, purify as before, and correlate wave-1 with wave-2
RT-IIV.

```{r stability, cache=TRUE}
long_sites <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota","pilot_uniandes_co_rural")
wv <- sc |> filter(dataset %in% long_sites, task_id %in% rt_tasks$task_id, !is.na(score)) |>
  group_by(user_id, task_id) |> arrange(age) |> mutate(wave = row_number(), nwave = max(wave)) |>
  filter(nwave >= 2, wave <= 2) |> ungroup() |> select(run_id, user_id, task_id, age, dataset, wave)

stab <- map_dfr(rt_tasks$task_id, function(tid){
  rw <- wv |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rw$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number)) |>
    mutate(lrt = log(rt_numeric))
  if (nrow(tr) < 50) return(NULL)
  nsite <- n_distinct(tr$dataset); form <- if (nsite>1) lrt ~ ns(trial_number,4)+dataset else lrt ~ ns(trial_number,4)
  tr$resid <- resid(lm(form, data = tr))
  pr <- tr |> group_by(run_id) |> filter(n() >= 8) |> summarise(iiv = sd(resid), .groups = "drop") |>
    left_join(rw, by = "run_id")
  w <- pr |> select(user_id, wave, iiv, age) |>
    pivot_wider(names_from = wave, values_from = c(iiv, age), names_sep = "") |>
    filter(!is.na(iiv1), !is.na(iiv2))
  sh <- tr |> filter(run_id %in% (rw |> filter(wave == 1) |> pull(run_id))) |>
    group_by(run_id) |> filter(n() >= 12) |> mutate(h = row_number() %% 2) |>
    group_by(run_id, h) |> summarise(v = sd(resid), .groups = "drop") |>
    pivot_wider(names_from = h, values_from = v, names_prefix = "h")
  rsh <- cor(sh$h0, sh$h1, use = "complete.obs")
  tibble(short = rt_tasks$short[rt_tasks$task_id == tid], n = nrow(w),
         gap_yr = median(w$age2 - w$age1), within_session = 2*rsh/(1+rsh),
         test_retest = cor(w$iiv1, w$iiv2))
}) |> filter(n >= 30)
```

```{r stability-table}
pooled <- map_dfr(rt_tasks$task_id, function(tid){
  rw <- wv |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rw$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number)) |> mutate(lrt = log(rt_numeric))
  if (nrow(tr) < 50) return(NULL)
  nsite <- n_distinct(tr$dataset); form <- if (nsite>1) lrt ~ ns(trial_number,4)+dataset else lrt ~ ns(trial_number,4); tr$resid <- resid(lm(form, data = tr))
  tr |> group_by(run_id) |> filter(n() >= 8) |> summarise(iiv = sd(resid), .groups = "drop") |>
    left_join(rw, by = "run_id") |> mutate(short = rt_tasks$short[rt_tasks$task_id == tid])
}) |> group_by(short) |> mutate(iivz = as.numeric(scale(iiv))) |> ungroup()
pw <- pooled |> select(user_id, short, wave, iivz) |> pivot_wider(names_from = wave, values_from = iivz, names_prefix = "w") |>
  filter(!is.na(w1), !is.na(w2))
pooled_r <- cor(pw$w1, pw$w2)
stab |> mutate(disattenuated = test_retest/within_session) |> arrange(desc(test_retest)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = glue::glue("Longitudinal test–retest stability of RT-IIV (Leipzig + Bogotá; pooled across tasks r = {round(pooled_r,3)}, n = {nrow(pw)})"))
```

```{r stability-plot, fig.height=4.3}
stab |> select(short, `within-session (split-half)` = within_session, `across-wave (test–retest)` = test_retest) |>
  pivot_longer(-short) |>
  ggplot(aes(value, reorder(short, value), color = name)) +
  geom_line(aes(group = short), color = "grey75") + geom_point(size = 2.6) +
  scale_color_manual(values = c("within-session (split-half)" = "#93a1a1", "across-wave (test–retest)" = "#268bd2"), name = NULL) +
  labs(x = "Reliability / stability", y = NULL,
       title = "RT-IIV is internally reliable but only moderately stable over time",
       subtitle = "Gap between within-session and across-wave = occasion-specific (state) variance") +
  theme(legend.position = "bottom")
```

RT-IIV is **partly trait-like, partly state**. Pooled across tasks the
test–retest correlation is ≈ 0.21, and even the best tasks — Hearts & Flowers
(0.47), Same-&-Different (0.40), ROAR-Word (0.37) — sit well below their
within-session reliabilities (0.6–0.92). The drop is the gap between the two
series in the plot: a large share of a child's RT-IIV on any given day is
**occasion-specific**, not a stable trait. Two things sharpen the read:

- The **speeded tasks** (H&F, ROAR-Word) that anchored the cross-task factor
  (§3) are again the most stable, reinforcing that the durable RT-IIV signal
  lives there.
- Even **disattenuated** for within-session unreliability, stability tops out
  around 0.4–0.5 — so the *true* RT-IIV trait still carries substantial
  year-to-year change (partly real developmental change: the speeded-task gaps
  span ~1.3 yr vs ~0.3–0.5 yr for the core tasks).

## Do the two-factor scores hold up better?

Pooling tasks into the §3 two-factor scores (speeded = swr/sre/hf;
untimed-cognitive = matrix/mrot/trog/math) *should* cancel task-specific noise
and stabilise the estimate — if the factors carry real shared variance. We score
each child's two RT-IIV factors per wave (mean of available within-task z-IIVs,
≥ 2 tasks) and test their wave-to-wave stability against the single tasks.

```{r factor-stability}
speeded <- c("swr","sre","hf"); cognitive <- c("matrix","mrot","trog","math")
fc <- pooled |>
  mutate(fac = case_when(short %in% speeded ~ "Speeded (swr/sre/hf)",
                         short %in% cognitive ~ "Cognitive (matrix/mrot/trog/math)", TRUE ~ NA_character_)) |>
  filter(!is.na(fac)) |> group_by(user_id, wave, fac) |>
  summarise(score = mean(iivz), ntask = n(), .groups = "drop") |> filter(ntask >= 2)
fac_stab <- fc |> select(user_id, wave, fac, score) |>
  pivot_wider(names_from = wave, values_from = score, names_prefix = "w") |>
  filter(!is.na(w1), !is.na(w2)) |> group_by(fac) |> summarise(n = n(), factor_score_r = cor(w1, w2), .groups = "drop")
contrast <- tibble(fac = c("Speeded (swr/sre/hf)","Cognitive (matrix/mrot/trog/math)"),
  mean_single_task_r = c(mean(stab$test_retest[stab$short %in% speeded]),
                         mean(stab$test_retest[stab$short %in% cognitive])),
  best_single_task_r = c(max(stab$test_retest[stab$short %in% speeded]),
                         max(stab$test_retest[stab$short %in% cognitive])))
fac_stab |> left_join(contrast, by = "fac") |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Two-factor RT-IIV factor-score stability vs single-task stability")
```

Aggregation **does not help** here. The cognitive factor score (test–retest
≈ 0.21, n = 304) only edges past the mean single cognitive task (≈ 0.18) and
matches the best single task; the speeded factor score (≈ 0.19, n = 57) is
actually *below* the best single speeded task (Hearts & Flowers alone, 0.47). The
reason follows directly from §3: the RT-IIV factors are **weak** (inter-task
r ≈ 0.1–0.17), so the shared variance the factor score targets is small, and
pooling mostly averages task-specific noise. The durable RT-IIV signal is
**task-specific** — carried by individual high-trial speeded tasks like H&F —
rather than by a shared RT-IIV factor.

The speeded factor's small n (57) is a **coverage** artifact, not missing data:
each speeded task individually has ~190–260 two-wave children, but the three are
repeated in *different* longitudinal subsamples — **H&F mainly in Leipzig**
(207 vs 52/0 in Bogotá urban/rural), **ROAR-Word/Sentence mainly in Bogotá**
(~165 vs ~22 in Leipzig). So only 57 children have ≥ 2 speeded tasks repeated
(511 have just one), and that composite is effectively the ROAR-Word + ROAR-Sentence
pair in Bogotá — which is why it underperforms H&F alone (a Leipzig task absent
from most of those 57). The four cognitive tasks are all core-battery tasks given
together, so they co-occur within child (n = 304).

# Synthesis

```{r save}
write_rds(list(summary = summ, cor_all = cmat, cor_reliable = cmatf,
               keep = keep, use_trials = USE_TRIALS,
               stability = stab, stability_pooled = pooled_r, factor_stability = fac_stab),
          here::here("data/rt_variability_07.rds"))
```

- **Purification helps, modestly.** Detrending the within-task time-course and
  accounting for autocorrelation roughly doubles cross-task RT-IIV coherence
  versus the naive per-run SD (≈ 0.05 → ≈ 0.10 across all tasks, ≈ 0.17 in the
  reliable subset). An age-varying trend does not matter (r ≈ 1); the trial-set
  choice has a modest per-task effect (r ≈ 0.74–0.98).
- **Trial count is decisive for stability.** RT-IIV reliability runs from ≈ 0.5
  (≤ ~25 trials) to ≈ 0.94 (120 trials). Short tasks cannot deliver a dependable
  per-child RT-IIV; designs that want this measure should ensure enough trials.
- **Coherence is real but narrow.** RT-IIV resolves into **two weak strands** — a
  speeded-task factor (ROAR-Word / Hearts & Flowers / ROAR-Sentence) and a weaker
  untimed-cognitive factor (TROG / vocab / math / matrix / mental rotation) —
  rather than a single general trait. Filtering to reliable, high-trial tasks
  leaves a passable single factor **dominated by the speeded tasks** (a plausible
  attentional-consistency signal).
- **Site and age differences are real.** RT-IIV declines with age (≈ −0.06 z/yr),
  the decline concentrated in the same speeded tasks that carry the signal. And
  children in **Bogotá are markedly more RT-variable** than Leipzig (urban +0.28,
  rural +0.33 SD, age-adjusted; Western ≈ Leipzig) — a large, consistent
  group-level shift (though site is only ~4% of total IIV variance) that may
  reflect administration medium, setting, or population and cautions against
  treating RT-IIV as directly cross-site comparable.
- **Only moderately stable over time.** Across ~1-year waves (Leipzig + Bogotá),
  RT-IIV test–retest is ≈ 0.21 pooled (up to 0.47 for H&F) — well below its
  within-session reliability, so much of a child's RT-IIV on a given day is
  occasion-specific. It is *partly* trait-like, and again most so for the speeded
  tasks. **Aggregating into the two-factor scores does not improve stability**
  (cognitive ≈ 0.21, speeded ≈ 0.19) — because the factors are weak, the durable
  RT-IIV signal is task-specific (esp. H&F), not shared-factor.
- **Implication.** If LEVANTE wants RT-IIV as an individual-difference measure,
  the leverage is in **speeded tasks with many trials**; reading it off untimed or
  short tasks will mostly capture task-specific strategy noise. This is a sharper,
  more optimistic read than 06's flat null — purification + trial-count awareness
  surfaces a real (if narrow) speeded-task RT-IIV factor — but its modest
  year-to-year stability means it is better treated as a partly-stable index than
  a fixed trait.
