---
title: "09 — LEVANTE network: enrollment ages & longitudinal coverage"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 7
execute:
  warning: false
  message: false
---

# Purpose

A planning view of the **whole LEVANTE network** — pilots, RfP1, and RfP2 — not
the analyzed data. We transcribe each funded site's main PI, data-collection
country, recruitment setting, and **target age range** from the Q1 2026 Advisory
Committee deck (`papers/Q1 LEVANTE Advisory committee meeting.pdf`), write it to a
CSV, and then visualize (1) the **enrollment ages** each study recruits and
(2) the **eventual age coverage** of the network after three waves.

```{r setup}
source(here::here("common.R"))
library(tidyverse)
```

# The site table

Transcribed from the deck's per-cohort tables and per-site annex cards (the
annex's explicit "Age range" is used as each site's target enrollment range).
`age_min`/`age_max` are enrollment ages; `cadence` is the planned wave spacing;
`school_based` marks school/classroom/daycare recruitment (vs. home, lab, online,
or community-cohort). See `notes` for deck inconsistencies.

```{r site-data}
sites <- tribble(
  ~cohort,       ~pi,                ~institution,                          ~country,                       ~age_min, ~age_max, ~school_based, ~cadence,    ~setting,            ~notes,
  "Pilots '23",  "Manuel Bohn",      "Leuphana University Lüneburg",        "Germany",                      5, 12, FALSE, "annual",   "online / family",   "",
  "Pilots '23",  "Julián Mariño",    "Universidad de los Andes",            "Colombia",                     5, 12, TRUE,  "annual",   "school",            "",
  "Pilots '23",  "Daniel Ansari",    "Western University",                  "Canada",                       5, 12, TRUE,  "annual",   "school",            "",
  "Pilots '23",  "Ben Domingue",     "Stanford University",                 "(secondary analysis)",         NA, NA, NA,    "n/a",      "secondary analysis","Analyzes other sites' pilot data; no recruitment.",
  "Pilots '25",  "Manuel Bohn",      "Leuphana University Lüneburg",        "Germany",                      2, 5,  FALSE, "annual",   "home / everyday",   "Vest-camera everyday recordings; 2 planned waves.",
  "Pilots '25",  "Charles Nelson",   "Boston Children's Hospital",          "United States",                2, 4,  FALSE, "annual",   "lab",               "",
  "Pilots '25",  "Tami Katzir",      "University of Haifa",                 "Israel",                       5, 10, TRUE,  "annual",   "school",            "Hebrew & Arabic cohorts.",
  "Pilots '25",  "Julián Mariño",    "Universidad de los Andes",            "Colombia",                     2, 4,  TRUE,  "annual",   "daycare",           "Early-childhood adaptation (2–4).",
  "Pilots '25",  "Ivy Nkrumah",      "University of Cape Coast",            "Ghana",                        5, 12, TRUE,  "annual",   "school",            "With C. Y. Kwaah; grades 1–5.",
  "RfP1",        "Alicia Forsberg",  "University of Sheffield",             "UK",                           6, 9,  TRUE,  "annual",   "school / online",   "Annex card lists 5–13; main table 6–9 (used).",
  "RfP1",        "Catherine Lebel",  "University of Calgary",               "Canada",                       0, 5,  FALSE, "annual",   "home",              "PdP cohort; LEVANTE added at 5–6. Annex lists U Alberta.",
  "RfP1",        "Rogier Kievit",    "Donders Institute (Radboud)",         "Netherlands",                  5, 9,  FALSE, "annual",   "lab / center",      "SMILEY (~5) & Skippy (~9) cohorts.",
  "RfP1",        "Zoe Ngo",          "Max Planck Inst. Human Development",  "Germany",                      4, 10, FALSE, "annual",   "lab",               "4- & 6-yr cohorts; wearable cameras + MRI.",
  "RfP1",        "Iliana Karipidis", "University of Zurich",                "Switzerland",                  5, 12, TRUE,  "annual",   "school",            "Mono- & multilingual cohorts.",
  "RfP1",        "Cecilia Calero",   "Universidad Torcuato Di Tella",       "Argentina",                    6, 9,  TRUE,  "annual",   "school",            "Cohorts 6–7 & 8–9 (annex card showed 6–7).",
  "RfP2",        "Diego Lucumí",     "Universidad de los Andes",            "Colombia",                     6, 11, TRUE,  "annual",   "school / community","Pacific-coast / PDET municipalities.",
  "RfP2",        "Dragan Gasevic",   "Monash University",                   "Colombia",                     9, 12, TRUE,  "biannual", "school",            "FLoRA SRL; data every 6 months.",
  "RfP2",        "Santiago Cueto",   "GRADE",                              "Peru",                         8, 11, TRUE,  "annual",   "school",            "APPrendemos cohort; 50 public schools.",
  "RfP2",        "Kwaku Poku Asante","Kintampo Health Research Center",     "Ghana",                        10, 12, FALSE, "annual",   "community cohort",  "GRAPHS birth-cohort follow-up.",
  "RfP2",        "Luciano Gasser",   "University of Teacher Education Bern", "Switzerland",                  9, 12, TRUE,  "annual",   "school",            "120 classrooms / 50 schools; CLASS.",
  "RfP2",        "Dana McCoy",       "Harvard University",                  "Peru",                         6, 8,  TRUE,  "annual",   "school / home",     "Rural Andean RCT follow-up.",
  "RfP2",        "Debarati Mukherjee","Public Health Foundation of India",  "India",                        6, 10, TRUE,  "annual",   "school / community","MAASTHI Bengaluru cohort.",
  "RfP2",        "Laura Di Giunta",  "Sapienza University of Rome",         "Colombia/Ecuador/Mexico/CR",   6, 12, TRUE,  "annual",   "school",            "Arukay; deck inconsistent re Costa Rica vs Dominican Republic."
) |>
  mutate(cohort = factor(cohort, c("Pilots '23","Pilots '25","RfP1","RfP2")),
         label  = paste0(pi, " · ", cohort))

write_csv(sites, here::here("09_site_coverage.csv"))
sites |> select(cohort, pi, institution, country, age_min, age_max, school_based, cadence, setting) |>
  knitr::kable(caption = "LEVANTE funded sites (23 grants across pilots, RfP1, RfP2)")
```

```{r quick-counts}
sites |> summarise(grants = n(),
                   recruiting = sum(!is.na(age_min)),
                   school_based = sum(school_based, na.rm = TRUE),
                   countries = n_distinct(country[!is.na(age_min)])) |>
  knitr::kable(caption = "Network at a glance")
```

The network spans **`r nrow(sites)` grants**; `r sum(!is.na(sites$age_min))`
recruit new participants (Stanford does secondary analysis only), of which
**`r sum(sites$school_based, na.rm=TRUE)`** use school / classroom / daycare
recruitment. Below we drop the secondary-analysis site from the age figures.

# 1. Enrollment ages, by study

A filled square means the study recruits children of that age (its target
`age_min`–`age_max`). Studies are grouped by cohort and ordered by youngest age.

```{r fig-enroll, fig.height=7}
ord <- sites |> filter(!is.na(age_min)) |> arrange(cohort, age_min, age_max) |> pull(label)
cohort_pal <- c("Pilots '23"="#b58900","Pilots '25"="#cb4b16","RfP1"="#268bd2","RfP2"="#2aa198")

enroll <- sites |> filter(!is.na(age_min)) |>
  reframe(age = 0:13, .by = c(label, cohort, age_min, age_max)) |>
  filter(age >= age_min, age <= age_max) |>
  mutate(label = factor(label, levels = rev(ord)))

ggplot(enroll, aes(age, label, fill = cohort)) +
  geom_tile(color = "white", width = 0.92, height = 0.82) +
  scale_fill_manual(values = cohort_pal, name = NULL) +
  scale_x_continuous(breaks = 0:13, expand = expansion(add = 0.5)) +
  labs(x = "Age (years)", y = NULL,
       title = "Enrollment ages recruited by each LEVANTE study",
       subtitle = "Filled = age recruited; colour = cohort") +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8))
```

The pilots and RfP1/RfP2 concentrate in the **school-age 5–12** band, with the
2025 pilots (Bohn, Nelson, Mariño) and RfP1 Lebel pushing **downward into early
childhood (0–5)** — matching the deck's stated RfP3 priorities (transition to
school; downward extension to 2–5).

# 2. Eventual coverage after three waves

For each recruiting study we extend its enrollment range by the follow-up its
cadence would add over **three waves** (annual → +2 years; biannual → +1 year),
giving the eventual age span the network would cover. Thick segment = enrollment
range; thin extension + open dot = ages reached by wave 3.

```{r fig-coverage, fig.height=7}
cov <- sites |> filter(!is.na(age_min)) |>
  mutate(step = if_else(cadence == "biannual", 0.5, 1),
         cov_max = age_max + 2 * step,
         label = factor(label, levels = rev(ord)))

ggplot(cov, aes(y = label, color = cohort)) +
  geom_segment(aes(x = age_max, xend = cov_max, yend = label),
               linewidth = 1, linetype = "dotted") +
  geom_segment(aes(x = age_min, xend = age_max, yend = label), linewidth = 2.6) +
  geom_point(aes(x = age_min), size = 2) +
  geom_point(aes(x = cov_max), size = 2, shape = 21, fill = "white", stroke = 1) +
  scale_color_manual(values = cohort_pal, name = NULL) +
  scale_x_continuous(breaks = 0:14, expand = expansion(add = 0.4)) +
  labs(x = "Age (years)", y = NULL,
       title = "Eventual longitudinal coverage after 3 waves",
       subtitle = "Solid = enrollment range; dotted = follow-up extension (●=baseline, ○=wave-3 age)") +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8))
```

```{r coverage-density, fig.height=3.2}
cov |> reframe(age = seq(min(age_min), max(cov_max)), .by = c(label, age_min, cov_max)) |>
  filter(age >= age_min, age <= cov_max) |>
  count(age, name = "n_studies") |>
  ggplot(aes(age, n_studies)) +
  geom_col(fill = "#268bd2") +
  scale_x_continuous(breaks = 0:14) +
  labs(x = "Age (years)", y = "# studies covering age",
       title = "Network age coverage (overlap) after 3 waves")
```

After three waves the network would cover roughly **ages 0–14**, with the
**densest overlap in the 6–12 school-age range** (where most pilots, RfP1, and
RfP2 sites pile up) and **thinner coverage at the early-childhood (0–4) and
older (13–14) ends** — the early end resting on just a few 2025-pilot / Lebel
studies. This is the empirical case for the RfP3 "downward extension" priority.

# Notes & caveats

- **Source:** the Q1 2026 Advisory Committee deck (gitignored under `papers/`).
  Age ranges are the annex cards' "Age range"; where a card conflicted with the
  cohort table we used the more specific value (Forsberg, Calero — see `notes`).
- **3-wave projection** is a *uniform planning assumption* (each site does three
  waves at its cadence); actual plans vary (e.g., Bohn-2025 plans two waves;
  Bern plans four). It illustrates eventual coverage, not committed design.
- **A few deck inconsistencies** are flagged in `notes` (Lebel: Calgary vs
  Alberta; Di Giunta: Costa Rica vs Dominican Republic as the 4th country).
- The CSV is written to `09_site_coverage.csv` for reuse.
