---
title: "09 — Choosing the RT-IIV purification pipeline: a systematic 2⁴ comparison"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

# The question

`07` purified RT before computing intra-individual variability, but made the
purification choices somewhat ad hoc. Here we choose them systematically. A
key idea (Mike): **item difficulty partially determines RT** — harder items take
longer — so on non-speeded tasks the difficulty *sequence* of administered items
injects RT variance that is not attentional inconsistency. Feeding the **fitted
item difficulties** back into the detrending model (e.g. `log(RT) ~
difficulty*trial`) should give a "difficulty-purified" IIV.

That gives four binary purification choices:

1. **detrend by trial number?** (within-task warm-up/fatigue)
2. **all trials vs correct only?**
3. **detrend by item difficulty?** (the new idea — uses the IRT `b` of each item)
4. **remove autocorrelation?** (AR(1) innovations)

We run **all 2⁴ = 16** combinations and score each on **split-half reliability**
(first vs second half — MSSD/AR are order-dependent) and **longitudinal
test–retest** (wave-1 vs wave-2, Leipzig + Bogotá), then pick the setting that
jointly maximizes the two.

```{r setup}
source(here::here("common.R"))
library(splines)
sc <- read_rds(here("data/scores_all_sites.rds"))
long_sites <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota","pilot_uniandes_co_rural")
ip <- load_item_parameters(); bylang <- c("vocab","theory-of-mind","swr","pa")
tasks <- tibble(
  task_id = c("egma-math","matrix-reasoning","mental-rotation","trog","vocab","same-different-selection",
              "hearts-and-flowers","theory-of-mind","swr","pa"),
  short   = c("math","matrix","mrot","trog","vocab","sds","hf","tom","swr","pa"),
  speeded = c(FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,TRUE,FALSE,TRUE,TRUE))   # difficulty-detrend should help the non-speeded ones
scor <- function(x, y){ ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 3 || sd(x[ok]) == 0 || sd(y[ok]) == 0) return(NA_real_); cor(x[ok], y[ok]) }
grid <- expand_grid(trial = c(FALSE,TRUE), trials = c("all","correct"), diff = c(FALSE,TRUE), ar = c(FALSE,TRUE))
```

# 1. Pipeline and evaluation

Each task's trials are loaded once (longitudinal sample), the fitted item
difficulty (`b`) joined per item, and each setting's IIV computed as the SD of
the purified log-RT residuals per run (AR(1) innovations when autocorrelation is
removed). The detrending model is built only from the chosen factors:

```{r run, cache=TRUE}
load_task <- function(tid){
  wv <- sc |> filter(dataset %in% long_sites, task_id == tid, !is.na(score)) |>
    group_by(user_id) |> arrange(age) |> mutate(wave = row_number(), nwave = max(wave)) |>
    filter(nwave >= 2, wave <= 2) |> ungroup() |> select(run_id, wave, language)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% wv$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000,
           !is.na(trial_number), !is.na(correct)) |>
    left_join(wv |> select(run_id, wave, language), by = "run_id") |> select(-any_of("difficulty"))
  ipx <- ip |> filter(task_id == tid)
  if (tid %in% bylang) tr <- tr |> inner_join(ipx |> transmute(language = subset, item_uid, difficulty), by = c("item_uid","language"))
  else tr <- tr |> inner_join(ipx |> transmute(item_uid, difficulty) |> distinct(item_uid, .keep_all = TRUE), by = "item_uid")
  tr |> mutate(lrt = log(rt_numeric))
}
eval_setting <- function(tr0, s){
  tr <- if (s$trials == "correct") filter(tr0, correct == 1) else tr0
  if (nrow(tr) < 150) return(c(rel = NA, retest = NA))
  terms <- c(); if (s$trial) terms <- c(terms, "ns(trial_number,3)"); if (s$diff) terms <- c(terms, "difficulty")
  if (s$trial && s$diff) terms <- c(terms, "difficulty:trial_number")
  form <- if (length(terms) == 0) lrt ~ 1 else as.formula(paste("lrt ~", paste(terms, collapse = "+")))
  tr$resid <- resid(lm(form, tr))
  tr <- tr |> group_by(run_id) |> arrange(trial_number, .by_group = TRUE) |>
    mutate(half = if_else(trial_number <= median(trial_number), 1L, 2L))
  if (s$ar) { tr <- tr |> mutate(lr = lag(resid)); rho <- scor(tr$resid, tr$lr); tr <- tr |> mutate(val = resid - rho*lr) }
  else tr <- tr |> mutate(val = resid)
  tr <- ungroup(tr)
  per <- function(by) tr |> group_by(across(all_of(by))) |> summarise(iiv = sd(val, na.rm = TRUE), n = n(), .groups = "drop") |> filter(n >= 8)
  rr <- per(c("run_id","half")) |> select(run_id, half, iiv) |> pivot_wider(names_from = half, values_from = iiv, names_prefix = "h")
  rel <- scor(rr$h1, rr$h2); rel <- if (is.na(rel)) NA else 2*rel/(1+rel)
  pr <- per("run_id") |> left_join(distinct(tr, run_id, user_id, wave), by = "run_id")
  rw <- pr |> select(user_id, wave, iiv) |> pivot_wider(names_from = wave, values_from = iiv, names_prefix = "w") |> filter(!is.na(w1), !is.na(w2))
  c(rel = rel, retest = scor(rw$w1, rw$w2))
}
res <- map_dfr(seq_len(nrow(tasks)), function(i){
  tr0 <- load_task(tasks$task_id[i])
  map_dfr(seq_len(nrow(grid)), function(j){
    m <- eval_setting(tr0, grid[j,])
    bind_cols(grid[j,], tibble(short = tasks$short[i], speeded = tasks$speeded[i], rel = m["rel"], retest = m["retest"])) })
})
```

# 2. All 16 settings

```{r table16}
summ <- res |> group_by(trial, trials, diff, ar) |>
  summarise(reliability = mean(rel, na.rm = TRUE), test_retest = mean(retest, na.rm = TRUE), .groups = "drop") |>
  mutate(joint = (reliability + test_retest)/2) |> arrange(desc(joint))
summ |> mutate(across(c(reliability, test_retest, joint), \(x) round(x, 3))) |>
  knitr::kable(caption = "All 16 purification settings, mean across tasks, sorted by joint (rel + test–retest)/2")
```

```{r scatter, fig.height=4.6}
summ |> mutate(`difficulty detrend` = diff, `trial set` = trials) |>
  ggplot(aes(test_retest, reliability, color = `difficulty detrend`, shape = `trial set`)) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(data = ~slice_max(.x, joint, n = 1),
                           aes(label = "best"), color = "black", size = 3.4, nudge_y = 0.01) +
  scale_color_manual(values = c(`FALSE` = "grey60", `TRUE` = "#268bd2")) +
  labs(x = "longitudinal test–retest", y = "split-half reliability",
       title = "Difficulty-detrending (blue) + all trials (circles) dominate",
       subtitle = "Each point one of the 16 settings; up-and-right is better") +
  theme(legend.position = "right")
```

The top settings all share **difficulty detrending** and **all trials**. The
single best by joint score is **trial + difficulty + all-trials + AR**, with the
AR-off version essentially tied.

# 3. What each choice buys (marginal effects)

```{r marginals}
map_dfr(c("trial","diff","ar","trials"), function(f){
  res |> group_by(level = as.character(.data[[f]])) |>
    summarise(reliability = mean(rel, na.rm = TRUE), test_retest = mean(retest, na.rm = TRUE), .groups = "drop") |>
    mutate(factor = f, .before = 1)
}) |> mutate(across(c(reliability, test_retest), \(x) round(x, 3))) |>
  knitr::kable(caption = "Marginal effect of each choice (averaging over the other three)")
```

- **All trials** is the biggest single win, on both metrics (consistent with 07).
- **Difficulty detrend helps** — it lifts reliability and test–retest, validating
  the hypothesis that part of RT variability is the item-difficulty sequence,
  not the child.
- **Trial detrend** helps modestly.
- **AR removal** is a genuine trade-off: it *raises* split-half reliability but
  *lowers* test–retest. The autocorrelation it strips out is partly a
  **person-stable** property, so removing it improves within-session consistency
  but discards longitudinally meaningful signal. For a trait measure, leave it in.

## Does difficulty-detrending help the non-speeded tasks specifically?

```{r diff-by-speeded}
res |> group_by(speeded = if_else(speeded, "speeded", "non-speeded"), diff) |>
  summarise(reliability = mean(rel, na.rm = TRUE), test_retest = mean(retest, na.rm = TRUE), .groups = "drop") |>
  mutate(diff = if_else(diff, "difficulty-detrended", "not"), across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Difficulty-detrending gains are largest for the non-speeded tasks")
```

As predicted, the difficulty-detrend gain is concentrated in the **non-speeded**
tasks (where RT is difficulty-driven); on the speeded tasks — where responses are
fast and difficulty-insensitive — it adds little.

## Per-task: 07's pipeline vs the recommended one

```{r per-task}
pick <- function(tr_, di) res |> filter(trial, trials == "all", diff == di, !ar) |> select(short, rel, retest)
cur <- pick(, FALSE) |> rename(rel_07 = rel, retest_07 = retest)
new <- pick(, TRUE)  |> rename(rel_new = rel, retest_new = retest)
cur |> left_join(new, by = "short") |> left_join(tasks |> select(short, speeded), by = "short") |>
  transmute(task = short, speeded, `Δ reliability` = rel_new - rel_07, `Δ test–retest` = retest_new - retest_07) |>
  arrange(desc(`Δ reliability`)) |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Adding difficulty-detrending (vs 07's trial-only): change in reliability & test–retest, per task")
```

# Recommendation

```{r save}
write_rds(list(settings = summ, per_task = res), here::here("data/rt_iiv_purification_09.rds"))
```

- **Adopt difficulty-detrending.** Across tasks it improves both reliability and
  longitudinal stability, with the gains where theory predicts — the non-speeded
  tasks whose RT is item-difficulty-driven. This directly addresses the
  difficulty confound behind 07's weak non-speeded RT-IIV.
- **Recommended pipeline:** `log(RT) ~ ns(trial) + difficulty + difficulty:trial`,
  on **all trials**, **keeping** the autocorrelation (do *not* AR-filter) — because
  AR removal trades away longitudinally-stable variance. This is the setting that
  best balances internal reliability and test–retest stability.
- **Caveat / scope.** The improvement is real but modest; it raises non-speeded
  RT-IIV without closing the gap to the speeded tasks, so 07's qualitative
  conclusion (the durable RT-IIV signal lives in speeded, high-trial tasks) is
  unchanged — but the non-speeded measure is now cleaner, and worth re-running 07
  on this pipeline if a single consistent definition is wanted.
