---
title: "Math (EGMA) — trial-level investigation of the apparent regression-to-mean in Bogotá"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Purpose

01 flagged that Bogotá math shows a striking pattern where high T1 scores
shrink toward the mean at T2 — Mike's "regression to the mean" question. We
test this at the trial level. **Key fact** uncovered while preparing this
notebook: every two-wave Bogotá math kid is non-adaptive at T1 and adaptive
at T2. So before we can claim "RTM," we have to rule out the much more
boring explanation that the CAT just changed the item set.

```{r setup}
source(here::here("common.R"))

scores <- read_rds(here::here("data", "scores_all_sites.rds"))
trials <- load_levante_trials(task_ids = "egma-math")
```

# Two-wave sample structure

```{r wave-coverage}
math_bog <- scores |>
  filter(dataset == "pilot_uniandes_co_bogota", task_id == "egma-math") |>
  group_by(user_id) |> arrange(age) |>
  mutate(wave = row_number()) |> ungroup() |>
  filter(wave <= 2)

math_bog |>
  group_by(user_id) |>
  filter(n_distinct(wave) == 2) |>
  ungroup() |>
  count(wave, adaptive, name = "n_runs") |>
  knitr::kable()
```

Every two-wave kid: T1 non-adaptive, T2 adaptive. The "RTM" diagnosis has to
contend with this confound up front.

# The pattern Mike spotted

```{r rtm-plot, fig.height=5}
pairs <- math_bog |>
  group_by(user_id) |>
  filter(n_distinct(wave) == 2) |>
  summarise(score_t1 = score[wave == 1],
            score_t2 = score[wave == 2],
            age_t1   = age[wave == 1],
            age_t2   = age[wave == 2],
            adapt_t1 = adaptive[wave == 1],
            adapt_t2 = adaptive[wave == 2],
            run_id_t1 = run_id[wave == 1],
            run_id_t2 = run_id[wave == 2],
            .groups  = "drop") |>
  mutate(t1_quartile = ntile(score_t1, 4) |>
                       factor(labels = c("Q1 (low)", "Q2", "Q3", "Q4 (high)")))

pairs |>
  ggplot(aes(x = score_t1, y = score_t2, color = t1_quartile)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_brewer(palette = "RdYlBu", direction = -1) +
  labs(x = "Score T1 (non-adaptive)", y = "Score T2 (adaptive)",
       title = "Math T1 vs T2 in Bogotá",
       subtitle = "Every kid: T1 non-adapt → T2 adapt. Note compression toward middle.",
       color = "T1 quartile")
```

```{r galton-numbers}
fit <- lm(score_t2 ~ score_t1, data = pairs)
broom::tidy(fit) |>
  knitr::kable(digits = 3,
               caption = "Galton's regression coefficients (T2 ~ T1)")
```

A slope below 1 (which we see) is the textbook RTM signature: high T1 scorers
move down at T2 and low T1 scorers move up. But here it could equally be
"the CAT scored everyone differently at T2." Let's separate the two.

# Trial-level: did the high-T1 kids see harder content at T2?

The math trial table doesn't carry a calibrated `difficulty` column (it's NA
for this task), so we use `item_group` as a proxy. Math items are grouped
into 12 sub-domains: addition, subtraction, multiplication, fractions,
number-line sliders, etc. The harder content (multiplication, fractions,
slider) lives at the top of the bank; counting and identification at the
bottom.

```{r item-group-dist, fig.height=5}
math_trials <- trials |>
  filter(dataset == "pilot_uniandes_co_bogota", task_id == "egma-math")

trials_with_quartile <- bind_rows(
  math_trials |> filter(run_id %in% pairs$run_id_t1) |>
    mutate(wave = "T1 (non-adapt)") |>
    inner_join(pairs |> select(run_id = run_id_t1, t1_quartile),
               by = "run_id"),
  math_trials |> filter(run_id %in% pairs$run_id_t2) |>
    mutate(wave = "T2 (adapt)") |>
    inner_join(pairs |> select(run_id = run_id_t2, t1_quartile),
               by = "run_id")
)

# Order item_groups from easy to hard (rough developmental ordering)
ig_order <- c("count", "ns_count", "identify", "ns_identify",
              "compare", "add", "subtract", "missing",
              "multiply", "fraction", "line", "slider")

trials_with_quartile |>
  mutate(item_group = factor(item_group, levels = ig_order)) |>
  count(t1_quartile, wave, item_group) |>
  group_by(t1_quartile, wave) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = item_group, y = pct, fill = wave)) +
  geom_col(position = "dodge") +
  facet_wrap(~t1_quartile, ncol = 4) +
  scale_fill_manual(values = c(`T1 (non-adapt)` = "#268bd2",
                                `T2 (adapt)`     = "#d33682")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Share of trials in item group",
       title = "Composition of items seen at each wave, by T1 quartile",
       subtitle = "Item groups ordered roughly easy→hard",
       fill = NULL) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r item-group-counts}
trials_with_quartile |>
  count(t1_quartile, wave, item_group) |>
  pivot_wider(names_from = wave, values_from = n, values_fill = 0L) |>
  arrange(t1_quartile, item_group) |>
  knitr::kable()
```

# Trial-level accuracy by wave and quartile

```{r accuracy, fig.height=5}
trials_with_quartile |>
  group_by(t1_quartile, wave) |>
  summarise(pct_correct = mean(correct, na.rm = TRUE),
            n_trials    = n(),
            n_kids      = n_distinct(user_id),
            .groups     = "drop") |>
  ggplot(aes(x = t1_quartile, y = pct_correct, fill = wave)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(`T1 (non-adapt)` = "#268bd2",
                                `T2 (adapt)`     = "#d33682")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "T1 quartile", y = "Trial-level accuracy",
       title = "Per-trial accuracy by wave",
       subtitle = "If T2 were harder for high-T1 kids, their accuracy would drop",
       fill = NULL) +
  theme(legend.position = "bottom")
```

# Item overlap between T1 and T2

If non-adaptive T1 and adaptive T2 sampled completely different items, then
the "RTM" effect can't be attributed to within-item performance change.

```{r item-overlap}
items_t1 <- math_trials |> filter(run_id %in% pairs$run_id_t1) |>
  pull(item_uid) |> unique() |> na.omit()
items_t2 <- math_trials |> filter(run_id %in% pairs$run_id_t2) |>
  pull(item_uid) |> unique() |> na.omit()

tibble(
  metric = c("Unique items in T1 (non-adapt)",
             "Unique items in T2 (adapt)",
             "Items in both",
             "T1-only", "T2-only"),
  value  = c(length(items_t1), length(items_t2),
             length(intersect(items_t1, items_t2)),
             length(setdiff(items_t1, items_t2)),
             length(setdiff(items_t2, items_t1)))
) |> knitr::kable()
```

# Within-kid: same items at both waves

For the subset of items that DO overlap, did the same kids answer them
consistently across waves? This is the cleanest within-individual change
estimate — no mode shift, no item shift.

```{r within-kid-item}
shared_items <- intersect(items_t1, items_t2)

t1_responses <- math_trials |>
  filter(run_id %in% pairs$run_id_t1, item_uid %in% shared_items) |>
  select(run_id, item_uid, correct_t1 = correct) |>
  inner_join(pairs |> select(user_id, run_id = run_id_t1),
             by = "run_id") |>
  select(user_id, item_uid, correct_t1)

t2_responses <- math_trials |>
  filter(run_id %in% pairs$run_id_t2, item_uid %in% shared_items) |>
  select(run_id, item_uid, correct_t2 = correct) |>
  inner_join(pairs |> select(user_id, run_id = run_id_t2),
             by = "run_id") |>
  select(user_id, item_uid, correct_t2)

same_items <- t1_responses |>
  inner_join(t2_responses, by = c("user_id", "item_uid")) |>
  inner_join(pairs |> select(user_id, t1_quartile, score_t1),
             by = "user_id")

cat("(kid, item) pairs answered at BOTH waves:", nrow(same_items), "\n")
cat("Unique kids with any overlap:", n_distinct(same_items$user_id), "\n")
cat("Median overlap items per kid:\n")
same_items |> count(user_id) |> pull(n) |> summary() |> print()
```

```{r same-item-acc, fig.height=4}
if (nrow(same_items) > 0) {
  same_items |>
    group_by(t1_quartile) |>
    summarise(
      n_obs        = n(),
      pct_t1       = mean(correct_t1, na.rm = TRUE),
      pct_t2       = mean(correct_t2, na.rm = TRUE),
      delta        = pct_t2 - pct_t1,
      .groups      = "drop"
    ) |>
    knitr::kable(digits = 3,
                 caption = "Accuracy on items seen at BOTH waves, by T1 quartile")

  same_items |>
    group_by(t1_quartile) |>
    summarise(pct_t1 = mean(correct_t1, na.rm = TRUE),
              pct_t2 = mean(correct_t2, na.rm = TRUE),
              .groups = "drop") |>
    pivot_longer(c(pct_t1, pct_t2), names_to = "wave", values_to = "pct") |>
    ggplot(aes(x = t1_quartile, y = pct, fill = wave)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = c(pct_t1 = "#268bd2", pct_t2 = "#d33682"),
                      labels = c("T1 (non-adapt)", "T2 (adapt)")) +
    scale_y_continuous(labels = scales::percent) +
    labs(x = "T1 quartile",
         y = "Accuracy on items seen at BOTH waves",
         title = "Within-item accuracy comparison",
         subtitle = "Cleaner than the global RTM plot — no item/mode confound",
         fill = NULL) +
    theme(legend.position = "bottom")
}
```

# Accuracy by item group and wave

If high-T1 kids are seeing harder content at T2, their per-item-group
accuracy on shared groups should stay similar (since each group spans a
narrow difficulty range), but the marginal mean shifts because the *mix* of
items shifted.

```{r acc-by-group, fig.height=5}
trials_with_quartile |>
  mutate(item_group = factor(item_group, levels = ig_order)) |>
  group_by(t1_quartile, wave, item_group) |>
  summarise(acc = mean(correct, na.rm = TRUE), n = n(), .groups = "drop") |>
  filter(n >= 20) |>
  ggplot(aes(x = item_group, y = acc, fill = wave)) +
  geom_col(position = "dodge") +
  facet_wrap(~t1_quartile, ncol = 4) +
  scale_fill_manual(values = c(`T1 (non-adapt)` = "#268bd2",
                                `T2 (adapt)`     = "#d33682")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Accuracy",
       title = "Per-item-group accuracy by wave",
       subtitle = "Within group, accuracy should be roughly stable if the kid hasn't actually changed",
       fill = NULL) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

# Verdict

```{r summary-stats}
slope <- coef(lm(score_t2 ~ score_t1, data = pairs))[2]

marginal <- trials_with_quartile |>
  group_by(t1_quartile, wave) |>
  summarise(acc = mean(correct, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = wave, values_from = acc,
              names_prefix = "acc_")

within_item <- if (exists("same_items") && nrow(same_items) > 0) {
  same_items |>
    group_by(t1_quartile) |>
    summarise(n = n(),
              pct_t1 = mean(correct_t1, na.rm = TRUE),
              pct_t2 = mean(correct_t2, na.rm = TRUE),
              delta  = pct_t2 - pct_t1,
              .groups = "drop")
} else NULL

marginal |> knitr::kable(digits = 3,
                          caption = "Marginal trial accuracy by wave (all items each kid saw)")
if (!is.null(within_item)) {
  within_item |> knitr::kable(digits = 3,
                               caption = "Within-item trial accuracy by wave (items each kid saw at BOTH waves)")
}
```

## Diagnosis

The plot in *§ The pattern Mike spotted* shows the classic RTM signature:
Galton slope ≈ 0.2, high-T1 kids drop ~2 θ at T2, low-T1 kids rise ~2 θ. But
the trial-level evidence shows this is **not within-individual regression to
the mean**; it's the CAT serving different items.

Three facts together make the case:

1. **Item composition shifts dramatically by quartile at T2.** For Q4 (high
   T1) kids, the CAT decided their ability warrants number-line problems
   ("line") — those go from ~5% of trials at T1 to ~64% at T2. The
   non-adaptive T1 form rarely tested them on this category. So their T2
   score is dominated by a category they hadn't been measured on before.

2. **Marginal trial accuracy looks like the score story.** Q4 kids drop from
   92% correct at T1 to 76% at T2; Q1 kids rise from 54% to 70%. But that's
   the marginal across very different item mixes.

3. **Within-item accuracy tells the opposite story.** On the 136 items
   shared between T1 and T2, *every* quartile is more accurate at T2:
   Q1 +11 pts, Q2 +5 pts, Q3 +7 pts, Q4 +7 pts. There is real growth in
   every quartile — the apparent decline at the top is purely the CAT
   routing.

## Implications for 02 / 03

- The Bogotá math T1 vs T2 IRT θ comparison cannot be interpreted as growth
  while every kid is NA→A. The within-item growth estimate (~+7%-points on
  shared items for high-T1 kids) is the more honest summary.
- For 02 / 03, treat Bogotá math two-wave θ deltas as **uninterpretable
  until** either (a) the CAT items get folded into a unified IRT calibration
  that's stable across modes, or (b) a fixed-form retest is added.
- The CAT routing decision is consistent — high-T1 kids really *are*
  measured as higher-ability by the CAT, which then shipped them
  developmentally appropriate harder content. This isn't a bug in the CAT;
  it's just incompatible with the naive longitudinal interpretation.

Sister problem worth flagging to the data coordinating center: T1's non-adapt
form sampled ~7% "slider" items and 0% "line"; T2's CAT sampled 0% "slider"
and lots of "line". These look like related number-line constructs. If they
were renamed between releases (rather than being genuinely different items),
the calibration won't recover from the relabel.
