---
title: "Shape Rotation (Mental Rotation) — the angle effect and item validity"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

# Question

Shape Rotation (mental rotation) asks the child to pick which of two
silhouettes matches a target when rotated. The canonical mental-rotation
signature is that difficulty (and RT) increases monotonically with the
**rotation angle**, and that 3-D figures are harder than 2-D. If the LEVANTE
item difficulties show this, it's strong evidence the task measures what it
should. Item structure: `mrot_{2d|3d}_{stimulus}_{angle}`.

```{r setup}
source(here::here("common.R"))
ip <- load_item_parameters() |>
  filter(task_id == "mental-rotation", subset == "all_items") |>
  mutate(dim   = str_extract(item_uid, "2d|3d"),
         stim  = str_extract(item_uid, "(?<=2d_|3d_)[a-z]+"),
         angle = as.integer(str_extract(item_uid, "[0-9]+$")))
```

# 1. The angle effect

```{r angle-plot, fig.height=5}
ip |>
  ggplot(aes(x = angle, y = difficulty, color = stim, group = stim)) +
  geom_hline(yintercept = 0, lty = 3, color = "grey70") +
  geom_line(aes(linetype = dim), linewidth = 0.9) +
  geom_point(size = 2.8) +
  scale_color_brewer(palette = "Set2") +
  scale_x_continuous(breaks = c(20,40,80,120,160)) +
  labs(x = "Rotation angle (degrees)", y = "Item difficulty (logits)",
       title = "Difficulty rises monotonically with rotation angle",
       subtitle = "The canonical mental-rotation signature; 3-D ≫ 2-D",
       color = "Stimulus", linetype = "Dimension")
```

```{r angle-slopes}
ip |> group_by(dim) |>
  summarise(n_items = n(),
            cor_angle_difficulty = cor(angle, difficulty),
            logits_per_45deg = coef(lm(difficulty ~ angle))[2] * 45,
            mean_discrimination = mean(discrimination),
            .groups = "drop") |>
  knitr::kable(digits = 3,
               caption = "Angle effect by dimension (difficulty near-linear in angle)")
```

The difficulty ordering is almost perfectly determined by rotation angle
(2-D r ≈ 0.99, 3-D r ≈ 0.88), increasing ~0.85 logits per 45° for 2-D figures.
3-D figures are much harder and shifted up. This is exactly the mental-rotation
effect — a clean validity result.

# 2. Accuracy by angle (the classic curve)

```{r acc-angle, fig.height=4.5}
tr <- load_levante_trials(task_ids = "mental-rotation") |>
  mutate(correct = as.numeric(correct),
         dim = str_extract(item_uid, "2d|3d"),
         angle = as.integer(str_extract(item_uid, "[0-9]+$")))
tr |> filter(!is.na(dim), !is.na(angle), item_uid %in% ip$item_uid) |>
  group_by(dim, angle) |>
  summarise(acc = mean(correct, na.rm = TRUE), n = n(), .groups = "drop") |>
  ggplot(aes(x = angle, y = acc, color = dim)) +
  geom_hline(yintercept = 0.5, lty = 3, color = "red") +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  scale_color_manual(values = c(`2d` = "#268bd2", `3d` = "#cb4b16")) +
  scale_x_continuous(breaks = c(20,40,80,120,160)) +
  labs(x = "Rotation angle (degrees)", y = "Accuracy",
       title = "Accuracy declines with rotation angle",
       subtitle = "Red = chance (2AFC). 3-D harder but above chance.",
       color = "Dimension")
```

# 3. Discrimination and the uncalibrated items

```{r disc}
ip |> arrange(dim, angle) |>
  select(item_uid, dim, stim, angle, difficulty, discrimination) |>
  knitr::kable(digits = 2, caption = "2PL item parameters")
```

```{r uncalibrated}
tr |> filter(!item_uid %in% ip$item_uid) |>
  count(item_uid, name = "n_trials") |> arrange(desc(n_trials)) |>
  knitr::kable(caption = "Items present in data but not calibrated (low-n polygon stimuli)")
```

Eight extra 2-D `polygon` stimuli appear in the data but weren't calibrated —
all very low-n (≤ 66 trials). They're a minor housekeeping item (exclude from
scoring), not a problem with the core duck/rabbit/3-D set. Discrimination is
healthy for 2-D (≈ 1.5–2.6, higher at larger angles) and lower for 3-D
(≈ 0.9–1.2), consistent with 3-D being harder and noisier.

# 4. Cross-site consistency

```{r site, fig.height=5}
acc <- tr |>
  filter(dataset %in% c("pilot_mpieva_de_main","pilot_uniandes_co_bogota"),
         item_uid %in% ip$item_uid) |>
  group_by(dataset_label, item_uid) |>
  summarise(acc = mean(correct, na.rm = TRUE), .groups = "drop") |>
  left_join(ip |> select(item_uid, dim, angle), by = "item_uid") |>
  pivot_wider(names_from = dataset_label, values_from = acc)
acc |>
  ggplot(aes(x = `Colombia — Bogotá`, y = `Germany — Leipzig`,
             color = dim, size = angle)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.85) +
  scale_color_manual(values = c(`2d` = "#268bd2", `3d` = "#cb4b16")) +
  coord_equal() +
  labs(x = "Item accuracy — Bogotá", y = "Item accuracy — Leipzig",
       title = glue::glue("Item difficulty replicates across sites (r = {round(cor(acc$`Colombia — Bogotá`, acc$`Germany — Leipzig`, use='complete.obs'),2)})"),
       color = "Dimension", size = "Angle")
```

# Verdict

- **Shape Rotation is a clean, valid measure.** Item difficulty shows the
  canonical mental-rotation **angle effect** (near-linear, r ≈ 0.99 for 2-D),
  with 3-D figures appropriately harder.
- 2PL discriminations are healthy; the model fits the design.
- Difficulty **replicates across sites** (r ≈ 0.96).
- Only housekeeping issue: a few low-n `polygon` stimuli are uncalibrated and
  should be excluded from scoring.
- No scoring concerns — usable as-is for growth/SEM.

```{r save}
write_rds(ip, here::here("data", "shape_rotation_item_params.rds"))
```
