---
title: "ROAR-Sentence (SRE) — scoring deepdive"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Scope

**Sentence Reading Efficiency (SRE)** is a speeded literacy measure adapted
from ROAR ([intro-sre](https://roar.stanford.edu/technical/intro-sre.html);
Yeatman et al., 2024). Children read simple sentences one at a time and judge
each *true* or *false* (2AFC), reading as many as they can in **180 seconds**.
ROAR's official score is **correct − incorrect** within the time window — the
subtraction is a guessing correction (a random guesser nets ≈ 0). LEVANTE
z-scores this count within dataset so it sits on a roughly comparable scale to
the IRT tasks (`score_type = "guessing_adjusted_number_correct_scaled"`).

**The scoring bug.** `levantemodels::score_sre()` computes
`filter(trial_number <= 180)` then `(sum(correct) - sum(!correct)) / 180` —
i.e. it truncates at the first 180 **trials**, not 180 **seconds**. (The
`/180` divisor is cosmetic: it cancels in the z-score.) This notebook asks:

1. **Timing forensics**: do runs actually last 180 seconds? (Spoiler: no —
   there is real structure here, including double-length runs.)
2. **Corrected scoring**: rate-normalize to a per-180s net-correct rate. How
   much do scores change? Do age / ROAR-Phoneme (PA) / ROAR-Word (SWR)
   correlations change, within and across sites?
3. **Efficiency scoring**: discussion of correctness+RT models
   (drift-diffusion IRT and alternatives) plus an EZ-diffusion proof of
   concept on the trial data.

Data: **`levante_data_pilots_hackathon:fpx0:v1_1`** via the new `levante`
package, cached by [`tasks/_pull_sre_hackathon.R`](\_pull_sre_hackathon.R).

```{r setup}
source(here::here("common.R"))

trials_file <- here("data/hackathon_v1_1_trials_roar.rds")
if (!file.exists(trials_file))
  stop("Run tasks/_pull_sre_hackathon.R first (requires Redivis auth).")

tr_all <- read_rds(trials_file)
sc_all <- read_rds(here("data/hackathon_v1_1_scores.rds"))

dataset_pal <- set_names(
  RColorBrewer::brewer.pal(4, "Set1"),
  dataset_labels_named[c("pilot_uniandes_co_bogota", "pilot_uniandes_co_rural",
                         "pilot_mpieva_de_main", "pilot_western_ca_main")])

label_dataset <- \(df) df |>
  mutate(dataset_label = factor(dataset_labels_named[dataset],
                                levels = names(dataset_pal)))

sre <- tr_all |>
  filter(task_id == "sre") |>
  mutate(correct = as.logical(correct),
         rt_s = rt_numeric / 1000,
         corpus = sub("_[0-9]+$", "", item_uid) |> str_remove("^sre_")) |>
  label_dataset()

sre_scores_pub <- sc_all |>
  filter(task_id == "sre", !is.na(score)) |>
  label_dataset()
```

```{r data-counts}
sre |>
  group_by(dataset_label) |>
  summarise(runs = n_distinct(run_id), children = n_distinct(user_id),
            trials = n(), .groups = "drop") |>
  knitr::kable(caption = "SRE trial data by dataset (hackathon v1_1)")
```

# 1. Timing forensics: is 180 seconds always 180 seconds?

Two per-run duration measures: **time-on-sentences** (cumulative RT — the
task records per-sentence RT in ms) and **wall-clock span** (last minus first
trial server timestamp). Neither is exactly the timer, but they bracket it:
cumulative RT omits inter-trial overhead; timestamp span omits the first
trial's duration and can inflate under server lag or pauses.

```{r durations}
run_time <- sre |>
  group_by(dataset, dataset_label, run_id, user_id) |>
  summarise(n_trials = n(),
            n_corpora = n_distinct(corpus),
            cum_rt = sum(rt_s),
            span = as.numeric(difftime(max(timestamp), min(timestamp),
                                       units = "secs")),
            n_correct = sum(correct), n_incorrect = sum(!correct),
            .groups = "drop")

run_time |>
  group_by(dataset_label) |>
  summarise(across(c(cum_rt, span),
                   list(q05 = ~quantile(.x, .05), median = median,
                        q95 = ~quantile(.x, .95), max = max)),
            .groups = "drop") |>
  knitr::kable(digits = 0,
               caption = "Per-run duration (seconds): cumulative RT and wall-clock span")
```

```{r duration-hist, fig.height=4}
run_time |>
  ggplot(aes(x = cum_rt, fill = dataset_label)) +
  facet_wrap(vars(dataset_label), scales = "free_y") +
  geom_histogram(binwidth = 10, boundary = 0, show.legend = FALSE) +
  geom_vline(xintercept = 180, lty = 2) +
  scale_fill_manual(values = dataset_pal) +
  labs(x = "Cumulative RT per run (s)", y = "Runs",
       title = "Most runs sum to just under 180 s of reading time",
       subtitle = "Dashed line = nominal 180 s deadline. Note the Canadian runs near 330 s.")
```

Three regularities:

- **The 180 s deadline is real but not exact.** Median time-on-sentences is
  ~160–165 s (the remainder is inter-trial overhead: wall-clock span medians
  ~180 s). So the timer fires on schedule for most runs.
- **A left tail of short runs exists everywhere** — Colombia especially (rural
  5th percentile ≈ 87 s of reading time). Under the current rule these
  short runs are scored as if the child had the full window.
- **Canada (Western) has genuinely double-length runs.**

## Block structure: some Canadian runs contain two 180-s blocks

The item bank prefix identifies sentence corpora. Colombia interleaves three
corpora (`spanish_lab`, `spanish_ai`, `spanish_ai-human`) *within one timed
block*; Germany uses one corpus (`uzh`). But 20 Canadian runs contain **two
sequential blocks** (`lab` sentences, then `ai` sentences), each with its own
~180-s window — roughly 360 s of testing scored as if it were one window.

```{r blocks}
run_blocks <- sre |>
  group_by(dataset_label, run_id, corpus) |>
  summarise(block_rt = sum(rt_s), n = n(), .groups = "drop_last") |>
  summarise(blocks = paste(sort(corpus), collapse = " + "),
            two_windows = sum(block_rt > 120) >= 2,
            .groups = "drop")

run_blocks |>
  count(dataset_label, blocks, two_windows) |>
  arrange(dataset_label, desc(n)) |>
  knitr::kable(caption = "Corpus composition of runs. two_windows = ≥2 corpora each with >120 s of reading time (sequential blocks, not interleaving)")
```

```{r trials-per-run, fig.height=4}
run_time |>
  ggplot(aes(x = n_trials, fill = dataset_label)) +
  facet_wrap(vars(dataset_label), scales = "free_y") +
  geom_histogram(binwidth = 10, boundary = 0, show.legend = FALSE) +
  geom_vline(xintercept = 180, lty = 2) +
  scale_fill_manual(values = dataset_pal) +
  labs(x = "Trials per run", y = "Runs",
       title = "Only Canada ever reaches the 180-trial truncation point",
       subtitle = "Dashed line = the current rule's truncation at 180 trials")
```

**Side-quest verdict:** the first-180-*trials* filter is a no-op for Germany
and Colombia (no run exceeds 180 trials — max 140). It bites only in Canada
(`r sum(run_time$n_trials > 180)` runs exceed 180 trials, max
`r max(run_time$n_trials)`), and there it truncates arbitrarily *across* the
two-block structure. Meanwhile the actual problem is the reverse: **effective
testing time varies from ~1 to ~4 minutes across runs**, and a raw count
treats them all as equal.

# 2. Corrected scoring: rate normalization

The corrected score is a **rate**: net correct per 180 s of reading time,

$$\text{score} = \frac{C - I}{\text{cumulative RT (s)}} \times 180,$$

then z-scored within dataset (preserving the pipeline's scaling convention).
We use time-on-sentences rather than wall-clock span as the denominator
because it is robust to server-timestamp lag and pauses; the two agree
closely (sensitivity check below). Runs with under 30 s of reading time are
excluded from rate scoring (extrapolation too aggressive).

First, verify we can reproduce the published (buggy) scores exactly:

```{r replicate-published}
old_score <- sre |>
  group_by(dataset, run_id) |>
  filter(trial_number <= 180) |>
  summarise(net = sum(correct) - sum(!correct), .groups = "drop") |>
  group_by(dataset) |>
  mutate(score_old = scale(net)[, 1]) |>
  ungroup()

old_score |>
  inner_join(sre_scores_pub, by = c("dataset", "run_id")) |>
  group_by(dataset_label) |>
  summarise(n = n(), r_with_published = cor(score_old, score), .groups = "drop") |>
  knitr::kable(digits = 3,
               caption = "Local reimplementation of the current rule vs published scores (r ≈ 1; residual differences come from a few scored runs with no rows in the trials table, which shift the z-scale slightly)")
```

```{r corrected-score}
min_time <- 30

new_score <- run_time |>
  filter(cum_rt >= min_time) |>
  mutate(rate = (n_correct - n_incorrect) / cum_rt * 180) |>
  group_by(dataset) |>
  mutate(score_new = scale(rate)[, 1]) |>
  ungroup()

scores_cmp <- new_score |>
  inner_join(old_score |> select(dataset, run_id, net, score_old),
             by = c("dataset", "run_id")) |>
  left_join(sre_scores_pub |> select(dataset, run_id, age),
            by = c("dataset", "run_id"))
```

Excluded by the 30-s floor: `r nrow(run_time) - nrow(new_score)` runs.

## A) How much do scores change?

```{r old-vs-new, fig.height=4.5}
cors <- scores_cmp |>
  group_by(dataset_label) |>
  summarise(n = n(), r = cor(score_old, score_new),
            r_spearman = cor(score_old, score_new, method = "spearman"),
            mean_abs_change = mean(abs(score_new - score_old)),
            max_abs_change = max(abs(score_new - score_old)),
            .groups = "drop")

knitr::kable(cors, digits = 2,
             caption = "Current (first-180-trials) vs corrected (rate per 180 s) z-scores")

scores_cmp |>
  ggplot(aes(x = score_old, y = score_new, color = dataset_label)) +
  facet_wrap(vars(dataset_label)) +
  geom_abline(lty = 3, color = "grey60") +
  geom_point(alpha = 0.4, size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = dataset_pal) +
  labs(x = "Current score (z)", y = "Corrected rate score (z)",
       title = "Current vs corrected SRE scores")
```

```{r movers}
scores_cmp |>
  mutate(change = score_new - score_old) |>
  slice_max(abs(change), n = 10) |>
  select(dataset_label, run_id, n_trials, cum_rt, n_correct, n_incorrect,
         score_old, score_new, change) |>
  knitr::kable(digits = 2, caption = "Largest score changes under rate normalization")
```

```{r change-vs-time, fig.height=4}
scores_cmp |>
  mutate(change = score_new - score_old) |>
  ggplot(aes(x = cum_rt, y = change, color = dataset_label)) +
  geom_hline(yintercept = 0, lty = 3) +
  geom_vline(xintercept = 180, lty = 2, color = "grey60") +
  geom_point(alpha = 0.5, size = 0.9) +
  scale_color_manual(values = dataset_pal, name = NULL) +
  labs(x = "Cumulative RT per run (s)", y = "Corrected − current (z)",
       title = "Who moves: short runs move up, double-length runs move down") +
  theme(legend.position = "bottom")
```

## B) Validity correlations, current vs corrected

Within each dataset: correlations of SRE with age, ROAR-Phoneme (θ), and
ROAR-Word (θ). Cross-run joins take each child's first run per task.

```{r validity}
first_run <- \(df) df |>
  group_by(dataset, user_id) |>
  arrange(time_started) |>
  slice(1) |>
  ungroup()

pa_swr <- sc_all |>
  filter(task_id %in% c("pa", "swr"), !is.na(score)) |>
  first_run() |>
  select(dataset, user_id, task_id, other = score) |>
  pivot_wider(names_from = task_id, values_from = other)

sre_first <- sre_scores_pub |>
  first_run() |>
  select(dataset, dataset_label, user_id, run_id, age) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_old, score_new),
             by = c("dataset", "run_id")) |>
  left_join(pa_swr, by = c("dataset", "user_id"))

validity <- sre_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_current = cor(score_old, crit),
            r_corrected = cor(score_new, crit), .groups = "drop") |>
  mutate(delta = r_corrected - r_current)

knitr::kable(validity, digits = 2,
             caption = "SRE validity correlations within dataset, current vs corrected scoring")
```

```{r validity-pooled}
pooled <- sre_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(criterion) |>
  summarise(n = n(),
            r_current = cor(score_old, crit),
            r_corrected = cor(score_new, crit), .groups = "drop") |>
  mutate(delta = r_corrected - r_current)

knitr::kable(pooled, digits = 2,
             caption = "Pooled across datasets (all scores z-scaled within dataset; PA/SWR θ are by-language, so pooled r mixes calibrations — interpret within-site values first)")
```

## Longitudinal stability (Bogotá, ~16-month gap)

Bogotá has ~180 children with SRE at two waves. Rank-order stability across
16 months is a demanding criterion: a better score should preserve more
true-score variance.

```{r stability}
stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(scores_cmp |> select(dataset, run_id, score_old, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap_days = as.numeric(difftime(time_started[2], time_started[1],
                                           units = "days")),
            across(c(score_old, score_new), list(t1 = ~.x[1], t2 = ~.x[2])),
            .groups = "drop") |>
  filter(gap_days > 300)

tibble(
  n_pairs = nrow(stab),
  median_gap_months = round(median(stab$gap_days) / 30.4, 1),
  stability_current = cor(stab$score_old_t1, stab$score_old_t2),
  stability_corrected = cor(stab$score_new_t1, stab$score_new_t2)
) |>
  knitr::kable(digits = 2, caption = "16-month rank-order stability, current vs corrected")
```

## Sensitivity: denominator choice

```{r denominator-sensitivity}
sens <- run_time |>
  filter(cum_rt >= min_time, span >= min_time) |>
  mutate(rate_rt = (n_correct - n_incorrect) / cum_rt * 180,
         rate_span = (n_correct - n_incorrect) / span * 180) |>
  group_by(dataset_label) |>
  summarise(n = n(), r_rt_vs_span = cor(rate_rt, rate_span), .groups = "drop")

knitr::kable(sens, digits = 3,
             caption = "Rate via cumulative RT vs via wall-clock span")
```

# 3. Fast guessing: the bridge to model-based scoring

The `C − I` correction neutralizes guessing *in expectation* (a guesser nets
≈ 0), but guessing still costs signal: guessed trials burn testing time and
add noise. And guessing is far from uniform across sites:

```{r fast-guessing, fig.height=4}
sre |>
  mutate(rt_bin = cut(rt_s, c(0, .5, 1, 2, 4, 8, 16, Inf))) |>
  group_by(dataset_label, rt_bin) |>
  summarise(accuracy = mean(correct), pct = n(), .groups = "drop_last") |>
  mutate(pct = pct / sum(pct)) |>
  ungroup() |>
  ggplot(aes(x = rt_bin, y = accuracy, color = dataset_label,
             group = dataset_label)) +
  geom_hline(yintercept = 0.5, lty = 3) +
  geom_line() +
  geom_point(aes(size = pct), alpha = 0.7) +
  scale_color_manual(values = dataset_pal, name = NULL) +
  scale_size_area(max_size = 6, guide = "none") +
  labs(x = "RT (s)", y = "Accuracy",
       title = "Sub-second responses are at chance — and their prevalence varies hugely by site",
       subtitle = "Point size ∝ share of site's trials in that RT bin; dotted line = 2AFC chance") +
  theme(legend.position = "bottom")
```

```{r guessing-table}
sre |>
  group_by(dataset_label) |>
  summarise(pct_rt_under_1s = mean(rt_s < 1),
            acc_under_1s = mean(correct[rt_s < 1]),
            acc_over_1s = mean(correct[rt_s >= 1]), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Fast-guessing prevalence by dataset")
```

This is a strong argument for scoring models that use the joint
(correct, RT) signal per trial, rather than counts alone.

# 4. Toward an efficiency score from correctness + RT

## The modeling landscape

SRE's design — self-paced 2AFC trials under a global deadline, with
per-trial RT and accuracy — admits several principled treatments:

1. **Rate normalization (implemented above).** Minimal fix: net correct per
   unit time. Already a speed–accuracy composite; requires no model. Its
   weakness: it treats a slow-and-accurate child and a fast-and-sloppy child
   with equal net rates as identical, and fast guessing adds noise it can
   only average over.

2. **Count models (Rasch Poisson counts / offset-Poisson GLMM).** The
   classical psychometrics of speed tests (Rasch, 1960): model the *count* of
   correct responses with a log link and `offset(log(time))`, giving each
   child a latent log-rate; corpus/site enter as fixed effects, children as
   random effects. This is rate normalization done as a measurement model —
   it handles varying time windows natively and yields standard errors. It
   still ignores per-trial RT structure.

3. **Diffusion IRT (van der Maas et al., 2011; Molenaar et al., 2015;
   `diffIRT`, or GLMM approximations).** Each trial's (choice, RT) is a
   Wiener first-passage process; person drift rate $v$ *is* reading
   efficiency, boundary separation $a$ is caution, $T_{er}$ is encoding/motor
   time. Items get drift/boundary loadings. This cleanly separates ability
   from speed–accuracy tradeoff and soaks up fast guesses (they are
   low-drift/low-boundary behavior, not ability). Cost: heavy estimation;
   item-level calibration needs coverage (fine in Colombia at ~60–80 runs per
   item; thin in Canada at ~9).

4. **Hierarchical speed–accuracy models (van der Linden, 2007).** Separate
   IRT model for accuracy and lognormal model for RT, with correlated person
   ability θ and speed τ. Less mechanistic than the diffusion model; scales
   well; efficiency is a composite of θ and τ.

5. **Signed residual time scoring (Maris & van der Maas, 2012).** Score
   $(2x - 1)(d - t)$ per item — correct-and-fast is best, wrong-and-fast is
   worst. Elegant and provably optimal under a specific IRT model, but it
   assumes a *per-item* deadline $d$; SRE has only a global deadline, so $d$
   would be an arbitrary pseudo-deadline. Not a natural fit.

## Zero-filled IRT (ROAR-team suggestion)

The ROAR team has begun exploring a simple item-based approach: **fill the
response matrix with zeros for unreached items, then fit a standard IRT
model**. Speed enters the score through the block of zeros — a fast accurate
reader clears more of the bank. We fit one dense Rasch model per item bank
(Spanish: Bogotá + rural jointly, 1,382 runs × 376 items; German: 151 × 200;
English: 91 × 530 after dropping 18 never-reached items;
[`_fit_sre_zerofill_irt.R`](\_fit_sre_zerofill_irt.R)).

```{r zerofill}
zf <- read_rds(here("data/sre_zerofill_scores.rds")) |>
  group_by(dataset) |>
  mutate(zf_z = scale(theta_zerofill)[, 1]) |>
  ungroup()

zf_first <- sre_first |>
  inner_join(zf |> select(dataset, run_id, zf_z), by = c("dataset", "run_id"))

zf_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_rate = cor(score_new, crit),
            r_zerofill = cor(zf_z, crit), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Zero-filled Rasch vs corrected rate score: validity")
```

```{r zerofill-stability}
zf_stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(zf |> select(dataset, run_id, zf_z), by = c("dataset", "run_id")) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            across(c(zf_z, score_new), list(t1 = ~.x[1], t2 = ~.x[2])),
            .groups = "drop") |>
  filter(gap > 300)

tibble(n_pairs = nrow(zf_stab),
       rate = cor(zf_stab$score_new_t1, zf_stab$score_new_t2),
       zerofill = cor(zf_stab$zf_z_t1, zf_stab$zf_z_t2)) |>
  knitr::kable(digits = 2,
               caption = "Bogotá 16-month stability: zero-filled Rasch vs corrected rate")
```

```{r zerofill-guessing}
guess_share <- sre |>
  group_by(dataset, run_id) |>
  summarise(pct_fast = mean(rt_s < 1), .groups = "drop")

zf_first |>
  inner_join(guess_share, by = c("dataset", "run_id")) |>
  group_by(dataset_label) |>
  summarise(n = n(),
            r_rate_fastguess = cor(score_new, pct_fast),
            r_zerofill_fastguess = cor(zf_z, pct_fast), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Correlation of each score with a run's fast-guess share (RT < 1 s). Positive = the score REWARDS guessing")
```

Where behavior is clean, zero-fill is essentially the rate score by another
route: r = .94 with the corrected rate in Germany, near-identical validity
there, and Bogotá 16-month stability a statistical tie (.72 vs .72). But it
has **no guessing correction**: a fast guesser converts wasted trials into
lucky 1s among the zeros, so the score correlates **positively** with a
run's fast-guess share (+.46 in rural Colombia) where the rate score
correlates negatively (−.42), and validity collapses exactly there (rural
PA r = .27 vs .57; rural age .19 vs .42).

The obvious repair — a fixed 2AFC guessing floor (Rasch with g = .5, as in
the LEVANTE multiple-choice models) — **backfires**: unreached items are
*structural* zeros, and whole-bank accuracy (~10% for a typical run) sits
far below any g = .5 floor, so the model cannot represent the data (EM
fails to converge; validity collapses everywhere, e.g. rural age r = .08).
The floor is right for "attempted and guessed," wrong for "never reached."
A repaired variant would need a different treatment of the two zero types —
at which point the model is no longer the simple suggestion, and the rate
score already does the job.

## Proof of concept: EZ-diffusion per run

The EZ-diffusion model (Wagenmakers, van der Maas, & Grasman, 2007) inverts
the simple diffusion model from three per-person summary statistics
(accuracy, mean and variance of correct RTs) — no item structure, no
iterative fitting. It is the cheapest possible test of whether
diffusion-style decomposition adds validity beyond the rate score.

```{r ez}
ez_diffusion <- function(pc, vrt, mrt, s = 0.1) {
  s2 <- s^2
  L <- qlogis(pc)
  x <- L * (L * pc^2 - L * pc + pc - 0.5) / vrt
  v <- sign(pc - 0.5) * s * x^(1 / 4)
  a <- s2 * L / v
  y <- -v * a / s2
  mdt <- (a / (2 * v)) * (1 - exp(y)) / (1 + exp(y))
  tibble(drift = v, boundary = a, ter = mrt - mdt)
}

ez_run <- sre |>
  filter(rt_s >= 0.25, rt_s <= 30) |>
  group_by(dataset, dataset_label, run_id, user_id) |>
  filter(n() >= 15, sum(correct) >= 10) |>
  summarise(n = n(), pc_raw = mean(correct),
            mrt = mean(rt_s[correct]), vrt = var(rt_s[correct]),
            .groups = "drop") |>
  # edge-correct accuracy away from 0.5 and 1
  mutate(pc = pmin(pmax(pc_raw, 0.5 + 1 / (2 * n)), 1 - 1 / (2 * n)))

ez_run <- bind_cols(ez_run, ez_diffusion(ez_run$pc, ez_run$vrt, ez_run$mrt))

ez_run |>
  group_by(dataset_label) |>
  summarise(n_runs = n(), across(c(drift, boundary, ter), median),
            .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Median EZ-diffusion parameters by dataset")
```

```{r ez-validity}
ez_first <- sre_first |>
  inner_join(ez_run |> select(dataset, run_id, drift, boundary, ter),
             by = c("dataset", "run_id"))

ez_validity <- ez_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_corrected_rate = cor(score_new, crit),
            r_drift = cor(drift, crit),
            r_boundary = cor(boundary, crit), .groups = "drop")

knitr::kable(ez_validity, digits = 2,
             caption = "EZ drift vs corrected rate score: validity correlations")
```

```{r ez-stability}
ez_stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(ez_run |> select(dataset, run_id, drift),
             by = c("dataset", "run_id")) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            drift_t1 = drift[1], drift_t2 = drift[2],
            rate_t1 = score_new[1], rate_t2 = score_new[2], .groups = "drop") |>
  filter(gap > 300)

tibble(n_pairs = nrow(ez_stab),
       stability_rate = cor(ez_stab$rate_t1, ez_stab$rate_t2),
       stability_drift = cor(ez_stab$drift_t1, ez_stab$drift_t2)) |>
  knitr::kable(digits = 2,
               caption = "Bogotá 16-month stability: corrected rate vs EZ drift")
```

```{r ez-scatter, fig.height=4.5}
ez_first |>
  ggplot(aes(x = score_new, y = drift, color = dataset_label)) +
  facet_wrap(vars(dataset_label), scales = "free_y") +
  geom_point(alpha = 0.4, size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = dataset_pal) +
  labs(x = "Corrected rate score (z)", y = "EZ drift rate v",
       title = "EZ drift vs the corrected rate score")
```

## Hierarchical speed–accuracy model (van der Linden 2007 style)

The next rung up from EZ: model accuracy and RT jointly but *separately*,
with crossed run × item random effects (fit by
[`_fit_sre_hierarchical_sa.R`](\_fit_sre_hierarchical_sa.R)):

- accuracy: `correct ~ 0 + dataset + (1|run_id) + (1|item_uid)` (logit GLMM)
  → run-level **ability θ**, with item easiness absorbed by item effects
  (the "IRT part" that EZ lacks);
- speed: `log(rt) ~ 0 + dataset + (1|run_id) + (1|item_uid)` (lognormal LMM)
  → run-level **speed τ** (sign-flipped RT effect), with item
  time-intensities absorbed.

An efficiency composite is z(θ) + z(τ), standardized within dataset.

```{r sa-load}
sa_file <- here("data/sre_sa_estimates.rds")
sa <- read_rds(sa_file)

sa_run <- sa$run |>
  group_by(dataset) |>
  mutate(theta_z = scale(theta)[, 1], tau_z = scale(tau)[, 1],
         eff = scale(theta_z + tau_z)[, 1]) |>
  ungroup()

sa_run |>
  label_dataset() |>
  group_by(dataset_label) |>
  summarise(n_runs = n(), r_theta_tau = cor(theta, tau), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Ability–speed correlation by dataset: near zero in DE/Canada, strongly NEGATIVE in rural Colombia — 'fast' there mostly means guessing")
```

```{r sa-validity}
sa_first <- sre_first |>
  inner_join(sa_run |> select(dataset, run_id, theta_z, tau_z, eff),
             by = c("dataset", "run_id"))

sa_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_rate = cor(score_new, crit),
            r_theta = cor(theta_z, crit),
            r_tau = cor(tau_z, crit),
            r_composite = cor(eff, crit), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Hierarchical SA model vs corrected rate score: validity")
```

```{r sa-stability}
sa_stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(sa_run |> select(dataset, run_id, theta_z, tau_z, eff),
             by = c("dataset", "run_id")) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            across(c(theta_z, tau_z, eff, score_new),
                   list(t1 = ~.x[1], t2 = ~.x[2])), .groups = "drop") |>
  filter(gap > 300)

tibble(n_pairs = nrow(sa_stab),
       rate = cor(sa_stab$score_new_t1, sa_stab$score_new_t2),
       theta = cor(sa_stab$theta_z_t1, sa_stab$theta_z_t2),
       tau = cor(sa_stab$tau_z_t1, sa_stab$tau_z_t2),
       composite = cor(sa_stab$eff_t1, sa_stab$eff_t2)) |>
  knitr::kable(digits = 2,
               caption = "Bogotá 16-month stability: SA-model scores vs corrected rate")
```

The decomposition is diagnostic gold but scoring lead. The ability–speed
correlation flips sign across sites (≈ 0 in DE/Canada, −.23 Bogotá, **−.52
rural CO**): where fast guessing is common, the "speed" dimension is
contaminated by disengagement, so neither τ nor any fixed θ+τ composite
travels across sites. The composite matches the rate score where behavior is
clean (r ≈ .9 in DE/Canada) and underperforms it where it isn't (rural age
validity .28 vs .42; Bogotá stability .68 vs .72).

## Diffusion IRT (pooled brms Wiener model)

The full model: every trial's (choice, RT) as a Wiener process, with
dataset fixed effects plus run *and item* random effects on drift (the IRT
part), run effects on boundary separation and non-decision time, bias fixed
at .5 (~64k trials; item banks are disjoint across sites, so sites are
linked by priors, not anchors). Prototyped locally with brms
([`_fit_sre_ddm.R`](\_fit_sre_ddm.R)), then exported to raw Stan
([`_export_sre_ddm_stan.R`](\_export_sre_ddm_stan.R)) and fit on Sherlock
([`sherlock_sre_ddm/`](sherlock_sre_ddm/); 24 cores, 9.5 h, 4 × 500/500).
Drift-side parameters converged acceptably (R̂ ≤ 1.06); the run-level
non-decision-time hierarchy mixed poorly (R̂ ≈ 1.6) — a known fragility of
ndt under a hard global deadline — which cautions against interpreting the
ndt estimates but leaves the drift comparison intact.

```{r ddm-load}
ddm_file <- here("data/sre_ddm_estimates.rds")
has_ddm <- file.exists(ddm_file)
```

```{r ddm-results, eval=has_ddm}
ddm <- read_rds(ddm_file)

ddm_run <- ddm$run |>
  group_by(dataset) |>
  mutate(drift_z = scale(drift_re)[, 1]) |>
  ungroup()

ddm_first <- sre_first |>
  inner_join(ddm_run |> select(dataset, run_id, drift_z, bs_re, ndt_re),
             by = c("dataset", "run_id"))

ddm_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_rate = cor(score_new, crit),
            r_drift = cor(drift_z, crit),
            r_boundary = cor(bs_re, crit), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Diffusion IRT drift vs corrected rate score: validity")
```

```{r ddm-stability, eval=has_ddm}
ddm_stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(ddm_run |> select(dataset, run_id, drift_z),
             by = c("dataset", "run_id")) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            across(c(drift_z, score_new), list(t1 = ~.x[1], t2 = ~.x[2])),
            .groups = "drop") |>
  filter(gap > 300)

tibble(n_pairs = nrow(ddm_stab),
       rate = cor(ddm_stab$score_new_t1, ddm_stab$score_new_t2),
       ddm_drift = cor(ddm_stab$drift_z_t1, ddm_stab$drift_z_t2)) |>
  knitr::kable(digits = 2,
               caption = "Bogotá 16-month stability: diffusion-IRT drift vs corrected rate")
```

```{r ddm-corr, eval=has_ddm, fig.height=4.5}
ddm_first |>
  ggplot(aes(x = score_new, y = drift_z, color = dataset_label)) +
  facet_wrap(vars(dataset_label)) +
  geom_point(alpha = 0.4, size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = dataset_pal) +
  labs(x = "Corrected rate score (z)", y = "DDM-IRT drift (z)",
       title = "Diffusion-IRT drift vs the corrected rate score")
```

The item side of the model behaves exactly as an IRT model should — item
drift effects track empirical item accuracy and (negatively) item RT:

```{r ddm-items, eval=has_ddm, fig.height=4}
item_emp <- sre |>
  group_by(item_uid) |>
  summarise(accuracy = mean(correct), median_rt = median(rt_s), n = n(),
            .groups = "drop") |>
  filter(n >= 20)

ddm$item |>
  inner_join(item_emp, by = "item_uid") |>
  ggplot(aes(x = drift_re, y = accuracy)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black") +
  labs(x = "Item drift effect (logit-ish)", y = "Empirical item accuracy",
       title = "The IRT part works: item drift effects track item difficulty")
```

So the item-level structure that EZ lacked is genuinely recovered here — and
person-level drift *still* underperforms the rate score where they differ.
That localizes the problem: it is not missing item information, it is that
the diffusion decomposition discards stable, valid between-child variance in
*time allocation* (caution, pacing) that the rate score keeps.

# Verdict

**1. The timer is mostly honest, but the trial-count rule was aimed at the
wrong problem.** The first-180-*trials* filter is a no-op in Germany and
Colombia (max 140 trials/run) and bites only in Canada. The real
heterogeneity is in *time*: 20 Canadian runs contain two sequential ~180-s
blocks (~360 s of testing scored as one window), and there is a left tail of
short runs everywhere (rural Colombia 5th percentile ≈ 87 s of reading time)
that the current rule scores as if they had the full window.

**2. Rate normalization changes little in aggregate, a lot for specific
kids.** Corrected-vs-current r ≈ .997 (DE), .975–.978 (CO), but **.82 in
Canada** (mean |Δ| = 0.45 z there; the double-block runs drop, short runs
rise). Validity correlations with age/PA/SWR move by |Δr| ≤ .03 everywhere
except Canadian age validity, which improves (.32 → .39); Bogotá 16-month
stability ticks up (.70 → .72). So the correction is about **fairness of
individual scores** (especially any longitudinal use of the Canadian data),
not about rescuing aggregate structure — which was already fine.

**3. Efficiency modeling: a four-rung ladder, and the rate score wins every
rung.** Every model-based alternative is less valid and/or less stable than
the corrected rate score where they differ. On the most demanding
criterion, Bogotá 16-month rank-order stability:

| score | stability | fatal flaw elsewhere |
|---|---|---|
| corrected rate (C − I per 180 s) | **.72** | — |
| zero-filled Rasch (ROAR suggestion) | .72 | rewards guessing (r = **+.46** with fast-guess share in rural CO vs −.42 for rate; PA validity .27 vs .57) |
| hierarchical SA composite (θ + τ) | .68 | speed dimension means different things by site |
| diffusion-IRT drift (full Wiener, item REs) | .56 | discards stable pacing/caution variance |
| EZ-diffusion drift (no items) | .44 | all of the above, plus no item structure |

Item structure is not the missing ingredient: the full diffusion-IRT fit
recovers sensible item parameters (item drift effects correlate .52 with
item accuracy, −.69 with item median RT) and item information lifts drift
stability from .44 (EZ) to .56 — but still far short of .72. Validity tells
the same story (e.g. Bogotá age r: rate .60, DDM drift .53; DE age: .54 vs
.44), with only scattered small wins for model-based scores (DE SWR: DDM
.62 vs rate .58). The diagnosis: SRE's net rate *is* the behaviorally
meaningful quantity — it bundles decision efficiency **and** stable
individual differences in pacing/caution, and every decomposition that
strips the latter out throws away true-score variance. The site-dependent
ability–speed correlation (≈ 0 in DE/Canada, −.52 in rural CO, where fast =
guessing) additionally means no fixed recombination of decomposed parts
travels across sites.

**Recommendations for the DCC:**

- **Adopt rate normalization now** (patch below). It fixes the Canadian
  double-block and short-run distortions at zero modeling cost and is what
  the ROAR "net correct in 3 minutes" score means.
- **If a measurement model is wanted**, the next step is an
  **offset-Poisson / Rasch-Poisson counts GLMM** (net-correct count with
  `offset(log(time))`, corpus fixed effects, child random effects) — it is
  rate normalization with standard errors and corpus adjustment, not a new
  scoring philosophy.
- **Diffusion IRT is now tested, not just argued against: don't use it for
  the primary SRE score.** The full hierarchical Wiener fit (9.5 h on
  Sherlock) yields drift scores that are less stable (.56 vs .72) and mostly
  less valid than the rate score. The decomposition (drift vs boundary vs
  ndt) remains available in `data/sre_ddm_estimates.rds` for substantive
  questions about speed–accuracy tradeoff and its development — boundary
  separation declines with age in every dataset — but that is a research
  variable, not a score.
- **Use RT for data quality regardless**: sub-second responses sit at chance
  and range from 3% of trials (DE) to 31% (rural CO). A fast-guess flag
  (e.g. proportion of RT < 1 s) is a cheap, useful run-level QC covariate.

Proposed patch to `levantemodels::score_sre()`:

```r
score_sre <- \(trial_data_task, dataset) {
  trial_data_task |>
    group_by(.data$run_id) |>
    summarise(elapsed = sum(.data$rt_numeric, na.rm = TRUE) / 1000,
              net = sum(.data$correct) - sum(!.data$correct)) |>
    filter(.data$elapsed >= 30) |>
    mutate(score = scale(.data$net / .data$elapsed * 180)[, 1],
           score_type = "guessing_adjusted_rate_scaled") |>
    select(-"elapsed", -"net")
}
```
